---
title: Minimizing fStress by Majorizing A Gauss-Newton Approximation
author: 
    - name: Jan de Leeuw
      orcid: 0000-0003-1420-1797
      email: jan@deleeuwpdx.net
      affiliation: 
        - name: University of California Los Angeles
          city: Los Angeles
          state: CA
          url: www.ucla.edu
      license: "CC0"
date: last-modified
date-format: long
bibliography: [mypubs.bib, total.bib]
number-sections: true
pdf-engine: lualatex
keep-tex: true
format:
   pdf:
    fontsize: 12pt
    include-in-header: 
     - preamble.tex
    keep-tex: true
    link-citations: true
    documentclass: scrartcl
    number-sections: true
   html:
    fontsize: 12pt
    include-in-header: 
     - preamble.css
    keep-md: true
    number-sections: true
toc: true
toc-depth: 3
editor: source
papersize: letter
graphics: true
link-citations: true
mainfont: Times New Roman
abstract: TBD
---

```{r loadpackages, echo = FALSE}
suppressPackageStartupMessages(library(knitr, quietly = TRUE))
suppressPackageStartupMessages(library(tinytex, quietly = TRUE))
suppressPackageStartupMessages(library(microbenchmark, quietly = TRUE))
suppressPackageStartupMessages(library(MASS, quietly = TRUE))
suppressPackageStartupMessages(library(isotone, quietly = TRUE))
```

```{r load_sources, echo = FALSE}
source("smacofPlots.R")
source("smacofSSFStressR.R")
source("smacofSSFStressC.R")
source("ekmanData.R")
source("smacofDataUtilities.R")
source("smacofPlots.R")
source("smacofTorgerson.R")
```
\sectionbreak 

**Note:** This is a working manuscript which will be expanded/updated
frequently. All suggestions for improvement are welcome. All Rmd, tex,
html, pdf, R, and C files are in the public domain. Attribution will be
appreciated, but is not required. The files can be found at
<https://github.com/deleeuw/rStress> 

\sectionbreak

# Loss Functions

The Multidimensional Scaling (MDS) loss function *stress*, originally proposed by @kruskal_64a, @kruskal_64b, is defined as
\begin{equation}
\sigma(x,\delta):=\frac12\sum_{k=1}^K w_k(\delta_k-d_k(x))^2,\label{eq-stressdef}.
\end{equation}
In definition \eqref{eq-stressdef} 

* $w$ is a vector of $K$ known positive *weights*;
* $\delta\in\Delta$, with $\Delta$ a set of $K$-element vectors with non-negative *dissimilarities*;
* $x:=\text{vec}(X)$, with $X$ the *configuration*, an $n\times p$ matrix of $n$ *points*
in $p$ *dimensions*;
* $d(x)$ are $K$ Euclidean *distances* between some or all pairs of points. 

Thus $K\leq\frac12n(n -1)$. For each $1\leq k\leq K$ there is a pair of indices $(i,j)$ with $1\leq i<j\leq n$
and a block-diagonal matrix $A_k$, which is the direct sum of $p$ copies of $(e_i-e_j)(e_i-e_j)'$,
where the $e_i$ are unit vectors (columns of the identity matrix of order $n$).
Thus $\smash{d_k(x)=\sqrt{x'A_kx}}$. 

In *metric* least squares MDS the set $\Delta$ is a singleton, and MDS minimizes stress over $x$. In *non-metric*
MDS the loss function is minimized over both $x$ and $\delta\in\Delta$. For the *ordinal* 
version of non-metric MDS, for example, $\Delta$ consists of all vectors with $0\leq\delta_1\leq\cdots\leq\delta_K$.

The Multidimensional Scaling (MDS) loss function *fStress* (@groenen_deleeuw_mathar_C_95) is a generalization of stress defined as
\begin{equation}
\sigma_f(x,\delta):=\frac12\sum_{k=1}^K w_k(\delta_k-f(d_k(x)))^2,\label{eq-fdef}
\end{equation}
with $f$ is a known real-valued function defined on $(0,+\infty)$. We assume that

* $f$ is strictly increasing on $(0,+\infty)$,
* $f$ is continuously differentiable on $(0,+\infty)$.

Thus we do not require $f$ to be defined at zero. We do *not* assume, but recommend,

* $f(0)=0,$
* $\lim_{x\rightarrow\infty}f(x)=\infty.$

If $f$ is the identity we have Kruskal's
stress, if $f$ is the square we have *sstress* used by @takane_young_deleeuw_A_77.
If $f$ is the logarithm we have the loss function used by @ramsay_77.

@groenen_deleeuw_mathar_C_95 give a slightly different
definition of fStress. They use
\begin{equation}
\sigma_f(x):=\frac12\sum_{k=1}^K w_k(f_r(\delta_k^0)-f(d_k(x)))^2,\label{eq-fdef2}
\end{equation}
where the $\delta_k^0$ are observed numerical dissimilarities. 
For all practical purposes definitions \eqref{eq-fdef} and \eqref{eq-fdef2} are
equivalent. In the metric case we simply use $\smash{\delta_k=f(\delta_k^0)}$
in \eqref{eq-fdef} by preprocessing. In the ordinal case $\delta_k^0$ and $f_r(\delta_k^0)$
are isotone, so the ordinal constraints in both definitions are the same.
For our purposes it is important that they also show that at a local minimum
of fStress we have $d_k(x)>0$ for all $k$ for which $w_k\delta_k>0$, generalizing
the result for stress proved by @deleeuw_A_84f.
No explicit algorithm to minimize fStress was given, but the formulas for the first 
and second derivatives in @groenen_deleeuw_mathar_C_95 can of course be used in general-purpose optimization routines.

An important special case of fStress is *rStress* (also known as *powerStress*), which is
\begin{equation}
\sigma_r(x):=\frac12\sum_{k=1}^K w_k(\delta_k-d_k^r(x))^2\label{eq-rdef}
\end{equation}
Both Kruskal's stress with $r=1$, sstress with $r=2$, and the limiting case logarithmic stress with $r\rightarrow 0$ are again special cases. There have been various attempts to extend the majorization (or MM) method for MDS (@deleeuw_C_77) to rStress. References and links to various unpublished reports are in @deleeuw_E_17r. The recent smacofx package (@rusch_deleeuw_mair_hornik_25) has R code for the rstressMin() function that implements the majorization method in @deleeuw_groenen_mair_E_16a.

In non-metric MDS we need to exclude the trivial solution
with $x=0$ and $\delta=0$. This is done by imposing the normalization constraint
\begin{equation}
\eta^2(\delta):=\frac12\sum_{k=1}^Kw_k\delta_k^2=1.\label{eq-normal}
\end{equation}
It is important to observe that for rStress we have homogeneity

# Algorithm

The technique proposed in this paper to minimize non-metric fStress 
is in the Alternating Least Squares (ALS) family. We start with an initial estimate
$x^{(0)}$. We then minimize $\sigma_f$ over $\delta$ in $\Delta\cap\Sigma$ for 
the current $\smash{d(x^{(0)})}$. The minimizer $\smash{\delta^{(0)}}$ is then used to minimize
fStress over $x$ for the current disparities, yielding $x^{(1)}$. These two steps are
alternated until $x$ and $\delta$ do not change any more. Starting
in iteration $\nu=0$ we compute
\begin{subequations}
\begin{align}
\delta^{(\nu)}&=\mathop{\text{argmin}}_{\delta\in\Delta\cap\Sigma}\sigma_f(x^{(\nu)},\delta),\label{eq-als1}\\
x^{(\nu+1)}&=\mathop{\text{argmin}}_x\sigma_f(x,\delta^{(\nu)})\label{eq-als2}.
\end{align}
\end{subequations}
We then increase $\nu$ by one and go into the next iteration. 
And so on, until convergence.


## Normalized Cone Regression

In metric MDS there is no need for the first step \eqref{eq-als1}, because $\delta^{(\nu)}$ is equal to
the normalized dissimilarities $\delta$ for all $\nu$.
In the non-metric case step \eqref{eq-als1} is needed, but compared to \eqref{eq-als2} it is comparatively easy. We have to compute the least squares
projection of $f(d(x))$ on the cone $\Delta$ and then normalize this projection
to give it length one. If $\Delta$ is the cone of monotone vectors, as in
ordinal MDS, the cone projection is *monotone regression*. The fact that
projection on the intersection of $\Delta$ and $\Sigma$ is the same thing as
normalizing the projection on $\Delta$ is due to @deleeuw_U_75a 
and more generally (and more rigorously) to @bauschke_bui_wang_18.
Since this step is the same as in standard MDS algorithms such as
smacof (@deleeuw_mair_A_09c, @mair_groenen_deleeuw_A_22) we do not go
into details here, and just refer to the literature.

## Majorization

The second step \eqref{eq-als2}, minimizing over $x$ for fixed current $\delta$, is more complicated. There is
no analytic solution, unlike the first step, and minimization requires an iterative process of its own. Thus, except in some very special cases, the second step requires an infinite number of "inner" iterations. Since implementing an infinite number of iterations is impossible
we have to truncate the inner iteration sequence at some point. In our software we deviate from the strict ALS framework by not minimizing fStress over $x$ for fixed $\delta$,
but by taking a single "inner" iteration and merely decrease fStress. If this is
done judiciously we still obtain a convergent sequence of updates. This is similar to the strategy in other
MDS algorithms such as smacof (@deleeuw_C_77) and alscal (@takane_young_deleeuw_A_77).

In smacof the inner iteration step is a majorization step, by now more commonly known as an MM step.
Briefly, we find a function $\kappa_f(x;y)$ such that 

* $\kappa_f(x,y)\leq\sigma_f(x)$ for all $x$ and $y$ in $\Xi$, and 
* $\kappa_f(x;y)=\sigma_f(x)$ if and only if $x=y$.

An inner iteration is of the form
\begin{equation}
x^{(\mu+1)}=\mathop{\text{argmin}}_{x\in\Xi}\kappa(x,x^{(\mu)}).\label{eq-inner}
\end{equation}
If $x^{(\mu+1)}=x^{(\mu)}$ we declare convergence and stop.
From \eqref{eq-inner},
\begin{equation}
\sigma_f(x^{(\mu+1)})\leq\kappa_f(x^{(\mu+1)},x^{(\mu)})<\kappa_f(x^{(\mu)},x^{(\mu)})=\sigma_f(x^{(\mu)}).
\label{eq-sandwich}
\end{equation}
Thus an inner majorization step upgrading $x$ for given $\delta$ decreases
fStress (unless we stop). Remember that in ALS the inner iterative process (indexed by $\mu$) goes on within step 2 of the "outer" update (indexed by $\nu$). 

In @deleeuw_groenen_mair_E_16a a majorization method was proposed to minimize rStress.
It majorizes $d^r$ as a function of $x$, and 
for this it uses the specific properties of the power function. It consequently cannot be
used for fStress. The technique is incorporated in the smacofx R package, and 
numerical experience indicates it works, but convergence can be painfully slow.
Because of the generality of the function $f$ (any differentiable increasing function)
it seems impossible to develop an majorization method for fStress along the same lines as
the one for rStress.
Consequently we go a somewhat different route in this paper. We do not
only deviate from the strict ALS procedure by only partially minimizing
loss over $x$ for fixed $d$, we also deviate from the strict majorization
approach by majorizing an approximation of fStress. 

In deriving the approximation and majorization we surpress the dependency of fStress
on $\delta$, because in the second ALS substep $\delta$ is just a vector of constants.
We first approximate $f(d(x))$ near $d(y)$ with 
\begin{equation}
f(d_k(x))\approx f(d_k(y))+\mathcal{D}f(d_k(y))(d_k(x)-d_k(y)).\label{eq-taylor}
\end{equation}
Define
\begin{equation}
\omega_f(x;y):=\sum_{k=1}^K w_k(f(\delta_k)-f(d_k(y))-\mathcal{D}f(d_k(y))(d_k(x)-d_k(y)))^2,\label{eq-defeta}
\end{equation}
Note that in \eqref{eq-taylor} the derivative is with respect to $d$, not with respect to $x$ or $y$. This
is a major difference with the majorizations in @deleeuw_groenen_mair_E_16a.

Note that $\omega_f(x;x)=\sigma_f(x)$ for all $x$ and if $f$ is the identity then 
$\omega_f(x;y)=\sigma_f(x)$ for all $x$ and $y$. If $f$ is linear with intercept 
$\beta$ and slope $\alpha$ then 
\begin{equation}
\omega_f(x,y)=a^2\sum_{k=1}^Kw_k(\delta_k-d_k(x))^2=a^2\sigma_r(x),\label{eq-linear}
\end{equation}
with $r=1$.

We now give an alternative and more convenient expression for $\omega_f(x,y)$. Define
\begin{subequations}
\begin{align}
\tilde w_k(y)&:=w_k\{\mathcal{D}f(d_k(y))\}^2,\label{eq-tildew}\\
\tilde\delta_k(y)&:=\frac{f(\delta_k)-f(d_k(y))}{\mathcal{D}f(d_k(y))}+d_k(y).\label{eq-tilded}
\end{align}
\end{subequations}
Then
\begin{equation}
\omega_f(x;y)=\sum_{k=1}^K\tilde w_k(y)(\tilde\delta_k(y)-d_k(x))^2.\label{eq-omegas}
\end{equation}
We see from \eqref{eq-tildew} that $\tilde w_k(y)$ is always non-negative and from \eqref{eq-tilded}
that $\tilde\delta_k(y)$ can be negative if $d_k(y)>\delta_k$. For convex $f$, however, we have
\begin{equation}
\tilde\delta_k(y)=\frac{f(\delta_k)-f(d_k(y))}{\mathcal{D}f(d_k(y))}+d_k(y)\geq d_k(x)\geq 0.\label{eq-convex}
\end{equation}

For rStress equations \eqref{eq-tildew} and \eqref{eq-tilded} become
\begin{subequations}
\begin{align}
\tilde w_k(y)&:=r^2w_kd_k^{2(r-1)}(y),\label{eq-tildewr}\\
\tilde\delta_k(y)&:=\frac{\delta_k^r+(r-1)d_k^r(y)}{rd_k^{r-1}(y)}.\label{eq-tildedr}
\end{align}
\end{subequations}
By convexity $\tilde\delta_k(y)$ is non-negative if $r\geq 1$. For $0<r<1$ we have 
$\tilde\delta_k(y)>0$ if and only if $\smash{d_k(y)<(1-r)^r\delta_k}$. If $\delta_k>d_k(y)$
then $\tilde\delta_k(y)>d_k(y)\geq 0$.

By writing $\omega_f(x,y)$ as in \eqref{eq-omegas} we are on familiar majorization terrain.
If $\delta_k(y)>0$ we use Cauchy-Schwartz, as in standard smacof,
\begin{subequations}
\begin{equation}
d_k(x)\geq\frac{1}{d_k(y)}\text{tr}\ x'A_ky,\label{eq-cs}
\end{equation}
and if $\delta_k(y)<0$ we use the arithmetic mean/geometric mean (AM/GM) inequality
in the form
\begin{equation}
d_k(x)\leq\frac12\frac{1}{d_k(y)}\{y'A_ky+x'A_kx\},\label{eq-amgm}
\end{equation}
\end{subequations}
Use of the AM/GM inequality for majorizing terms with negative dissimilarities 
was pioneered by @heiser_91.

Define the matrices
\begin{subequations}
\begin{align}
V(y)&:=\sum_{k=1}^K\tilde w_k(y)A_k,\\
B(y)&:=\sum_{\tilde\delta_k(y)>0}\tilde w_k\frac{\tilde\delta_k(y)}{d_k(y)}A_k,\\
H(y)&:=\sum_{\tilde\delta_k(y)<0}\tilde w_k\frac{|\tilde\delta_k(y)|}{d_k(y)}A_k.
\end{align}
\end{subequations}
Note that all three matrices (more precisely matrix-valued functions) are positive semi-definite.
Now, using these matrices and the inequalities \eqref{eq-cs} and \eqref{eq-amgm},
\begin{equation}
\omega_f(x,y)\leq\kappa(x,y):=\gamma(y)+x'(V(y)+H(y))x-2x'B(y)y,
\label{eq-matmajx}
\end{equation}
with
\begin{equation}
\gamma(y):=\sum_{k=1}^K\tilde w_k(y)\tilde\delta_k^2(y)+y'H(y)y
\end{equation}
which does not depend on $x$.



We now update using
$$
x^{(\mu+1)}=\mathop{\text{argmin}}_x\kappa_f(x;x^{(\mu)})=(V(x^{(\mu)})+H(x^{(\mu)}))^+B(x^{(\mu)})x^{(\mu)},
$$
which results in
$$
\sigma_f(x^{(\mu+1)})\approx\omega_f(x^{(\mu+1)};x^{(\mu)})\leq\kappa_f(x^{(\mu+1)};x^{(\mu)})<\kappa_f(x^{(\mu)};x^{(\mu)})=\sigma_f(x^{(\mu)}).
$$
Unlike in smacof there is no guarantee that $\sigma_f$ decreases from one iteration to the other. Monotone
convergence of loss function values will only happen if the approximation is good, i.e. if
$d(x^{(\mu+1)})$ is close to $d_k(x^{(\mu)})$. This requires closer monitoring of convergence than 
in smacof. Also note that if $\tilde\delta_k(y)<0$ for all $k$ then $B(y)=0$, which means that the update 
from $y$ is equal to zero. Again, this eventuality must be monitored.

# Software

The C function smacofSSRStressC() in the file smacofSSFStressC.c does either metric or ordinal MDS minimizing fStress for an arbitrary $f$. The C code is compiled into a shared library smacofSSFStress.so, which is loaded by the R front-end smacofSSFStressC.R

smacofSSRStressC()

## Data

For fStress.R the input is a matrix delta of dissimilarities. For smacofSSRStress.R the data and weights are in the MDS data format defined in @deleeuw_E_25b. This is a list with two scalars nobj and ndat, the number of objects and the number of data points (object pairs), and five vectors iind, jind, delta, blocks, weights  of length ndat. Vectors iind and jind have integers between 1 and n with iind > jind that code for which pairs of objects we have dissimilarity information. The dissimilarities are in delta, the weights in weights. Dissimilarities are sorted in increasing order and the blocks vectors codes the tie-blocks in the dissimilarities.



## Initial Estimate

There are two possibilities. If xinit is not NULL then it is used as the 
initial estimate of the configuration. This is useful, for example,
if we start the iterations for, say, $r=\tfrac12$ with the solition for $r=1$.
If xinit is NULL then the function smacofSSRStressInit() computes an initial estimate using classical MDS (also known as Torgerson-Gower MDS) on the dissimilarities.

First we make sure the weights add up to one. If there are
missing dissimilarities they are first imputed by setting them equal to the 
weighted mean of the non-missing dissimilarities. Then we double center the squared
dissimilarities and compute the classical scaling solution using
the eigs_sym() function from the RSpectra package (@qiu_mei_24).

Now we scale the solution, which is done with the R function
smacofSSRStressScale(). We first normalize the powered dissimilarities
such that their weighted sum of squares is equal to one. We then
choose $\lambda$ to minimize
$$
\sigma(\lambda):=\sum_k w_k(\delta_k^r-d_k^r(\lambda x))^2=\sum_k w_k(\delta_k^r-\lambda^rd_k^r(x))^2,
$$
where $x$ is the Torgerson-Gower solution. The minimum is attained for
$$
\hat\lambda:=\left[\frac{\sum_k w_k\delta_k^rd_k^r(x)}{\sum_kw_kd_k^{2r}(x)}\right]^{1/r}.
$$
Our initial estimate is $\hat\lambda x$ with distances $\hat\lambda d_k(x)$. This scaling is done also if xinit is not NULL.

## Functions

Some examples of $f$ are in the following table.

```{r kabf, echo = FALSE, eval = TRUE}
x <- matrix(c("log(rx + 1)", "r/(rx+1)", "(exp(y) - 1) / r", 
              "x^r", "rx^{r - 1}", "y^{1 / r}",
              "exp(rx) - 1", "r exp(rx)", "log(y + 1)/r"), 3, 3, byrow = TRUE)

rn <- c("log", 
       "power",
       "exponent")
row.names(x) <- rn
cn <- c("f_r(x)", "f'_r(x)", "f_r^{-1}(y)")
colnames(x) <- cn
kable(x, format = "simple", align = NULL, row.names = TRUE, caption = "f examples")
```



smacofSSFStressSelect.R

## Main

The arguments of smacofSSFStressR() and smacofSSFStressR() are
```{r theargsf, echo = FALSE, eval = FALSE}
args(fStress)
```

Thus fStress() needs an initial configuration. Parameters $f$ and $g$ are functions that compute the transformation $f$ and its derivative. The file
fStress.R includes flog, glog, fexp, gexp, and fPower, gPower. Parameter
$r$ is included in case $f$ and $g$ need additional arguments (as in the case of 
fPower and gPower). Since all code is in R, it is easy to add additional 
transformation functions if needed. The output of fStress() is 
```{r outputf, eval = FALSE}
list(
    delt = delta, 
    dhat = delf, 
    wmat = wght,
    conf = xnew,
    dist = dnew,
    loss = snew,
    itel = itel
  )
```

The arguments of smacofSSRStress() are
```{r theargsr, echo = FALSE, eval = FALSE}
args(smacofSSRStress)
```
The C code follows the conventions of the smacof versions in @deleeuw_E_25b. 
There is only a single parameter $rpow$ for the transformation. The output
of smacofSSRStress() is
```{r output, eval = FALSE}
list(
    delt = theData$delta,
    dhat = h$dhat,
    dist = h$edis,
    xmat = matrix(h$xnew, nobj, ndim),
    wmat = h$wght,
    loss = h$snew,
    ndim = ndim,
    init = xinit,
    itel = h$itel,
    nobj = nobj,
    iind = h$iind,
    jind = h$jind,
    weighted = weighted,
    ordi = ordinal,
    ties = h$ties
  )
```



## Plots

The four standard plots that come with the recent smacof programs () are

* smacofConfigurationPlot
* smacofShepardPlot
* smacofDistDhatPlot
* smacofResidualPlot

The output list of smacofSSFStress() can be fed immediately into these four plot routines.
They only have side effects (i.e. make a plot) and return NULL.

It is easy to see what smacofConfigurationPlot() does. It plots a two-dimensional
projection of the configuration. Its arguments are self-explanatory.
```{r eval = FALSE}
function(h, main = "ConfigurationPlot", labels = NULL, dim1 = 1, 
    dim2 = 2, pch = 16, col = "RED", cex = 1) 
```
A Shepard plot (@deleeuw_mair_C_15) traditionally plots dissimilarities on the
horizontal axis, and both disparities and distances on the vertical axis.
The $(\delta_k^0,\delta_k)$ points are connected by a straight line through the origin in the metric case and by an increasing step function in the ordinal case. The $(delta,dist)$ points
are optionally connected with vertical "fitlines" to the corresponding $(delta,dhat)$ points.
```{r eval = FALSE}
function(h, main = "ShepardPlot", fitlines = TRUE, colline = "RED", 
    colpoint = "BLUE", type = 0, lwd = 2, cex = 1, pch = 16) 
```
In the fStress context there is an extra arguments "type".
Remember that the list h has both a "what" and an "rpow" element.
The "type" argument is either one or zero. If it is zero (the default) then
we plot $f(\delta_k)$ on the horizontal axis and $\hat d_k$ and $f(d_k(x))$ on the vertical axis. If "type" is equal to one then we plot $\delta_k$ on the horizontal axis and $f^{-1}(\hat d_k)$ and $d_k(x)$ on the vertical axis.

smacofDistDhatPlot() has arguments
```{r eval = FALSE}
function(h, fitlines = TRUE, colline = "RED", colpoint = "BLUE", 
    main = "Dist-Dhat Plot", type = 0, cex = 1, lwd = 2, pch = 16) 
```
The "type" parameter works in the same way as in the Shepard plot.

smacofResidualPlot() is more complicated. It has arguments
```{r eval = FALSE}
function (h, main = "ResidualPlot", probs = seq(0, 1, 0.25), 
    dats = 1, whatplot = "ecdf", type = 0, qlines = TRUE, colpoints = "RED", 
    collines = "BLUE", lwdpoints = 2, lwdlines = 2, cex = 1, 
    pch = 16) 
```

As in the previous plots *type* computes residuals as 

*whatplot* can have the values *ecdf* (default) or *density*. R functions

*dats* chooses to plot the residual (default), its absolute value, or its square

# Examples

We analyse the colour data from @ekman_54 with different functions $f$, using both 
smacofSSFStressR() and using smacofSSFStressC().

```{r rfstressruns, echo = FALSE, cache = FALSE}
hl <- smacofSSFStressR(ekmanData, what = 1, verbose = FALSE)
hp1 <- smacofSSFStressR(ekmanData, rpow = .1, verbose = FALSE)
hp5 <- smacofSSFStressR(ekmanData, rpow = .5, verbose = FALSE)
h1 <- smacofSSFStressR(ekmanData, rpow = 1, verbose = FALSE)
h2 <- smacofSSFStressR(ekmanData, rpow = 2, verbose = FALSE)
h5 <- smacofSSFStressR(ekmanData, rpow = 5, verbose = FALSE)
he <- smacofSSFStressR(ekmanData, what = 2, verbose = FALSE)
```
```{r kableF, echo = FALSE}
floss <- c(hl$loss, hp1$loss, hp5$loss, h1$loss,h2$loss,h5$loss,he$loss)
fiter <- c(hl$itel, hp1$itel, hp5$itel, h1$itel,h2$itel,h5$itel,he$itel)
x <- matrix(c(fiter, floss), 7, 2)
rnames <- c("logarithm", 
       "power .1",
       "power .5",
       "power 1",
       "power 2", 
       "power 5",
       "exponent"
)
row.names(x) <- rnames
cnames <- c("itel", "fStress")
colnames(x) <- cnames
kable(x, format = "simple", align = "r", row.names = TRUE, caption = "smacofSSFStressR() results Ekman Data")
```

```{r kableR, echo = FALSE}
rloss <- c(0.005791783, 0.000273273, 0.005954311, 0.01721325,0.03288071,0.02031532,0.02073349)
riter <-  c(206, 854, 237, 35, 50,807, 32)
x <- matrix(c(riter, rloss), 7, 2)
rnames <- c("logarithm", 
       "power .1",
       "power .5",
       "power 1",
       "power 2", 
       "power 5",
       "exponent"
)
row.names(x) <- rnames
cnames <- c("itel", "fStress")
colnames(x) <- cnames
kable(x, format = "simple", align = "r", row.names = TRUE, caption = "smacofSSFStressC() results Ekman Data, metric MDS")
```

We see that for the power transforms the R and C functions arrive at the
same solution. The number of iterations is not exactly the same, because
of differences in the initial configuration. With smacofSSRStress() we can also
do ordinal MDS.


```{r kableRO, echo = FALSE}
rlosso <- c(0.0002466493, 6.567816e-06, 0.0001651214, 0.0005337258, 0.0009014502, 0.001100227, 0.0007894837)
ritero <-  c(336, 975, 207, 141, 233, 2638, 160)
x <- matrix(c(ritero, rlosso), 7, 2)
rnames <- c(
       "logarithm",
       "power .1",
       "power .5",
       "power 1",
       "power 2", 
       "power 5",
       "exponential"
)
row.names(x) <- rnames
cnames <- c("itel", "fStress")
colnames(x) <- cnames
kable(x, format = "simple", align = "r", row.names = TRUE, caption = "
      smacofSSFStressC() results Ekman Data, ordinal MDS")
```

The package microbenchmark (@mersmann_24) can be used to compare the
R and C versions. For $r=1$ the C version is about ten times faster,
for $r=.1$ and $r=5$ it is more like 30 times faster.

```{r kablemicrometric, echo = FALSE}
rtimes <- c(24030, 101871, 27919, 3967, 5929, 95857, 3883)
ctimes <-  c(690, 3115, 1073, 382, 474, 2920, 349)
x <- matrix(c(rtimes, ctimes), 7, 2)
rnames <- c(
       "logarithm",
       "power .1",
       "power .5",
       "power 1",
       "power 2", 
       "power 5",
       "exponential"
)
row.names(x) <- rnames
cnames <- c("R times", "C times")
colnames(x) <- cnames
kable(x, format = "simple", align = "r", row.names = TRUE, caption = "
      R and C timings Ekman Data in microseconds, metric MDS")
```
```{r eval = FALSE, echo = FALSE}
Unit: microseconds
                                                     expr       min          lq        mean
   smacofSSFStressR(ekmanData, what = 1, verbose = FALSE) 23038.638  23619.8540  28644.5889
   smacofSSFStressC(ekmanData, what = 1, verbose = FALSE)   618.198    668.6690    738.0201
 smacofSSFStressR(ekmanData, rpow = 0.1, verbose = FALSE) 99130.046 100679.3745 111889.0709
 smacofSSFStressC(ekmanData, rpow = 0.1, verbose = FALSE)  2901.857   3071.4125   3152.8094
 smacofSSFStressR(ekmanData, rpow = 0.5, verbose = FALSE) 26666.523  27385.2325  31396.8271
 smacofSSFStressC(ekmanData, rpow = 0.5, verbose = FALSE)  1011.757   1058.1485   1090.9260
   smacofSSFStressR(ekmanData, rpow = 1, verbose = FALSE)  3676.839   3922.1010   5143.4389
   smacofSSFStressC(ekmanData, rpow = 1, verbose = FALSE)   340.464    361.7635    401.6942
   smacofSSFStressR(ekmanData, rpow = 2, verbose = FALSE)  5583.913   5854.9640   6903.7317
   smacofSSFStressC(ekmanData, rpow = 2, verbose = FALSE)   407.048    438.7000    481.1440
   smacofSSFStressR(ekmanData, rpow = 5, verbose = FALSE) 92626.995  95049.3980 100948.4161
   smacofSSFStressC(ekmanData, rpow = 5, verbose = FALSE)  2747.779   2863.9730   2987.3613
   smacofSSFStressR(ekmanData, what = 2, verbose = FALSE)  3689.795   3823.3525   4600.4087
   smacofSSFStressC(ekmanData, what = 2, verbose = FALSE)   306.270    335.2160    406.1378
      median          uq        max neval    cld
  24030.0180  27291.9780  86524.760   100 a     
    690.3580    734.2280   4413.978   100  bc   
 101870.7320 105006.2890 220077.627   100    d  
   3115.0980   3169.8125   6647.986   100   c e 
  27918.6220  31085.1340  90309.224   100 a     
   1073.4005   1118.4800   1236.847   100  bc   
   3966.5040   4026.5485  65376.304   100  b  e 
    381.6485    439.1715    510.163   100   c   
   5929.2355   6030.7515  68230.150   100     e 
    474.4315    523.5905    611.720   100   c   
  95857.4055  97766.0990 158889.965   100      f
   2919.5690   2962.3525   6655.858   100   c e 
   3882.6795   3933.7860  65264.415   100   c e 
    349.4020    405.1620   4051.948   100   c   
```
```{r kablemicrononmetric, echo = FALSE}
rtimes <- c(111.93, 420.03, 115.90, 58.00, 95.26, 
            1172.62, 63.57)
ctimes <-  c(2.83, 8.96, 2.15, 1.46, 2.38, 23.53, 1.49)
x <- matrix(c(rtimes, ctimes), 7, 2)
rnames <- c(
       "logarithm",
       "power .1",
       "power .5",
       "power 1",
       "power 2", 
       "power 5",
       "exponential"
)
row.names(x) <- rnames
cnames <- c("R times", "C times")
colnames(x) <- cnames
kable(x, format = "simple", align = "r", row.names = TRUE, caption = "
      R and C timings Ekman Data in milliseconds, nonmetric MDS")
```
```{r eval = FALSE, echo = FALSE}
Unit: milliseconds
                                                                     expr         min          lq
   smacofSSFStressR(ekmanData, what = 1, verbose = FALSE, ordinal = TRUE)  106.306973  110.755452
   smacofSSFStressC(ekmanData, what = 1, verbose = FALSE, ordinal = TRUE)    2.542246    2.768279
 smacofSSFStressR(ekmanData, rpow = 0.1, verbose = FALSE, ordinal = TRUE)  402.058095  414.841136
 smacofSSFStressC(ekmanData, rpow = 0.1, verbose = FALSE, ordinal = TRUE)    8.338334    8.780068
 smacofSSFStressR(ekmanData, rpow = 0.5, verbose = FALSE, ordinal = TRUE)  109.666185  114.417306
 smacofSSFStressC(ekmanData, rpow = 0.5, verbose = FALSE, ordinal = TRUE)    1.954347    2.092128
   smacofSSFStressR(ekmanData, rpow = 1, verbose = FALSE, ordinal = TRUE)   55.808298   57.330997
   smacofSSFStressC(ekmanData, rpow = 1, verbose = FALSE, ordinal = TRUE)    1.311385    1.401995
   smacofSSFStressR(ekmanData, rpow = 2, verbose = FALSE, ordinal = TRUE)   89.000996   93.787131
   smacofSSFStressC(ekmanData, rpow = 2, verbose = FALSE, ordinal = TRUE)    2.181077    2.311272
   smacofSSFStressR(ekmanData, rpow = 5, verbose = FALSE, ordinal = TRUE) 1118.418623 1155.259296
   smacofSSFStressC(ekmanData, rpow = 5, verbose = FALSE, ordinal = TRUE)   22.259269   22.983288
   smacofSSFStressR(ekmanData, what = 2, verbose = FALSE, ordinal = TRUE)   59.894727   62.936353
   smacofSSFStressC(ekmanData, what = 2, verbose = FALSE, ordinal = TRUE)    1.383791    1.457017
        mean      median          uq         max neval       cld
  113.852865  111.933280  113.670327  176.549526   100 a        
    2.911093    2.826417    2.886216    6.084523   100  b       
  423.109740  420.029461  424.704014  487.229322   100   c      
    8.936843    8.955302    9.134964    9.351895   100    d     
  117.196493  115.901117  117.756203  179.792503   100 a        
    2.184620    2.150245    2.207522    5.484734   100  b       
   58.943635   57.998190   58.850642  122.780691   100     e    
    1.450922    1.456792    1.499903    1.597770   100  b       
   95.140413   95.257411   96.381037   99.670098   100      f   
    2.402289    2.382243    2.436856    5.536845   100  b       
 1176.369676 1172.620849 1186.844917 1265.874795   100       g  
   23.475012   23.533077   23.939449   24.703853   100        h 
   63.974907   63.568348   64.972679   68.236505   100         i
    1.497407    1.492666    1.539263    1.626183   100  b       
```

\sectionbreak
```{r chp1c, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Metric, r = .1", main ="", eval = FALSE}
par(pty="s")
smacofConfigurationPlot(chp1, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r chp1s, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Metric, r = .1", main = "", eval = FALSE}
smacofShepardPlot(chp1, rpow = .1, cex = 2)
```
\sectionbreak
```{r chp5c, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Metric, r = .5", main ="", eval = FALSE}
par(pty="s")
smacofConfigurationPlot(chp5, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r chp5s, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Metric, r = .5", main = "", eval = FALSE}
smacofShepardPlot(chp5, rpow = .5, cex = 2)
```
\sectionbreak
```{r ch1c, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Metric, r = 1", main ="", eval = FALSE}
par(pty="s")
smacofConfigurationPlot(ch1, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r ch1s, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Metric, r = 1", main = "", eval = FALSE}
smacofShepardPlot(ch1, rpow = 1, cex = 2)
```
\sectionbreak
```{r ch2c, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Metric, r = 2", main ="", eval = FALSE}
par(pty="s")
smacofConfigurationPlot(ch2, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r ch2s, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Metric, r = 2", main = "", eval = FALSE}
smacofShepardPlot(ch2, rpow = 2, cex = 2)
```
\sectionbreak
```{r ch5c, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Metric, r = 5", main ="", eval = FALSE}
par(pty = "s")
smacofConfigurationPlot(ch5, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r ch5s, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Metric, r = 5", main = "", eval = FALSE}
smacofShepardPlot(ch5, rpow = 5, cex = 2)
```

\sectionbreak
```{r chp1oc, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Non-metric, r = .1", main ="", eval = FALSE}
par(pty="s")
smacofConfigurationPlot(chp1o, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r chp1os, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Non-metric, r = .1", main = "", eval = FALSE}
smacofShepardPlot(chp1o, rpow = 2, cex = 2)
```
\sectionbreak
```{r chp5oc, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Non-metric, r = .5", main ="", eval = FALSE}
par(pty="s")
smacofConfigurationPlot(chp5o, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r chp5os, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Non-metric, r = .5", main = "", eval = FALSE}
smacofShepardPlot(chp5o, rpow = .5, cex = 2)
```
\sectionbreak
```{r ch1oc, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Non-metric, r = 1", main ="", eval = FALSE}
par(pty = "s")
smacofConfigurationPlot(ch1o, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r ch1os, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Non-metric, r = 1", main = "", eval = FALSE}
smacofShepardPlot(ch1o, rpow = 1, cex = 2)
```
\sectionbreak
```{r ch2oc, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Non-metric, r = 2", main ="", eval = FALSE}
par(pty = "s")
smacofConfigurationPlot(ch2o, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r ch2os, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Non-metric, r = 2", main = "", eval = FALSE}
smacofShepardPlot(ch2o, rpow = 2, cex = 2)
```
\sectionbreak
```{r ch5oc, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Non-metric, r = 5", main ="", eval = FALSE}
par(pty = "s")
smacofConfigurationPlot(ch5o, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r ch5os, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Non-metric, r = 5", main = "", eval = FALSE}
smacofShepardPlot(ch5o, rpow = 5, cex = 2)
```


\sectionbreak

# References