---
title: Minimizing fStress by Majorizing A Gauss-Newton Approximation
author: 
    - name: Jan de Leeuw
      orcid: 0000-0003-1420-1797
      email: jan@deleeuwpdx.net
      affiliation: 
        - name: University of California Los Angeles
          city: Los Angeles
          state: CA
          url: www.ucla.edu
      license: "CC0"
date: last-modified
date-format: long
bibliography: [mypubs.bib, total.bib]
number-sections: true
pdf-engine: lualatex
keep-tex: true
format:
   pdf:
    fontsize: 12pt
    include-in-header: 
     - preamble.tex
    keep-tex: true
    link-citations: true
    documentclass: scrartcl
    number-sections: true
   html:
    fontsize: 12pt
    include-in-header: 
     - preamble.css
    keep-md: true
    number-sections: true
toc: true
toc-depth: 3
editor: source
papersize: letter
graphics: true
link-citations: true
mainfont: Times New Roman
abstract: TBD
---

```{r loadpackages, echo = FALSE}
suppressPackageStartupMessages(library(knitr, quietly = TRUE))
suppressPackageStartupMessages(library(tinytex, quietly = TRUE))
suppressPackageStartupMessages(library(microbenchmark, quietly = TRUE))
suppressPackageStartupMessages(library(MASS, quietly = TRUE))
suppressPackageStartupMessages(library(isotone, quietly = TRUE))
```
```{r load_shared_library, echo = FALSE}
dyn.load("smacofSSFStress.so")
``` 

```{r load_sources, echo = FALSE}
source("smacofPlots.R")
source("smacofSSFStressR.R")
source("smacofSSFStressC.R")
source("ekmanData.R")
source("smacofDataUtilities.R")
source("smacofPlots.R")
source("smacofTorgerson.R")
```
\sectionbreak 

**Note:** This is a working manuscript which will be expanded/updated
frequently. All suggestions for improvement are welcome. All Rmd, tex,
html, pdf, R, and C files are in the public domain. Attribution will be
appreciated, but is not required. The files can be found at
<https://github.com/deleeuw/rStress> 

\sectionbreak

# Loss Functions

Metric Multidimensional Scaling (MDS) minimizes the loss function *(raw) stress*, originally proposed by @kruskal_64a, @kruskal_64b. Stress is
\begin{equation}
\sigma(x):=\sum_{k=1}^K w_k(\delta_k-d_k(x))^2,\label{eq-stressdef}.
\end{equation}
In definition \eqref{eq-stressdef} 

* $w$ is a vector of $K$ known positive *weights*;
* $\delta$ is a vector of $K$ known non-negative *dissimilarities*;
* $x:=\text{vec}(X)$, with $X$ the *configuration*, an $n\times p$ matrix of $n$ *points* in $p$ *dimensions*;
* $d(x)$ are $K$ Euclidean *distances* between some or all pairs of points. 

Thus $K\leq\frac12n(n -1)$. Our notation is slighty non-standard, because we
use vectors in our definition, and not matrices. For each $1\leq k\leq K$ there is a pair of indices $(i(k),j(k))$ with $1\leq i(k)<j(k)\leq n$. The unit vectors
$e_{i(k)}$ and $e_{j(k)}$ are used to define the matrix $A_k$, the direct sum of $p$ copies of $(e_{i(k)}-e_{j(k)})(e_{i(k)}-e_{j(k)})'$.
Suppose $n=4$ and $p=2$ and $(i(k),j(k))$ is $(2,4)$. Then
\begin{equation}
A_k=\begin{array}{rrrrrrrr}
0&0&0&0&0&0&0&0\\
0&+1&0&-1&0&0&0&0\\
0&0&0&0&0&0&0&0\\
0&-1&0&+1&0&0&0&0\\
0&0&0&0&0&0&0&0\\
0&0&0&0&0&+1&0&-1\\
0&0&0&0&0&0&0&0\\
0&0&0&0&0&-1&0&+1
\end{array}.
\end{equation}
The $A_k$, together with $x=\text{vec}(X)$, allow us to conveniently write $\smash{d_k(x)=\sqrt{x'A_kx}}$.

In non-metric MDS stress is a function of both $x$ and $\delta$, and we
minimize over both arguments. We require that $\delta\in\Delta$, with $\Delta$ is subset of the non-negative orthant of $\mathbb{R}^K$. In ordinal MDS, the
most common case, $\Delta$ is defined as the cone of vectors with $0\leq\delta_1\leq\cdots\leq\delta_k$,
together with some normalization requirement to exclude the trivial solution
with both $x$ and $\delta$ (and thus stress) equal to zero. The simplest such requirement is that $\delta$ is on the sphere $\mathbb{S}$ defined as the vectors $\delta$ for which
\begin{equation}
\sum_{k=1}^Kw_k^{\ }\delta_k^2=1.\label{eq-sphere}
\end{equation}

## Metric fStress

@groenen_deleeuw_mathar_C_95 propose a generalization of stress called *fStress*. It is defined as
\begin{equation}
\sigma_f(x):=\sum_{k=1}^K w_k(f(\delta_k)-f(d_k(x)))^2,\label{eq-fdef1}
\end{equation}
with $f$ a known real-valued function defined on $(0,+\infty)$. We assume that

1. $f$ is strictly increasing on $(0,+\infty)$,
2. $f$ is continuously differentiable on $(0,+\infty)$.

Thus we do *not* require $f$ to be defined at zero. We also do *not* require, but we have a preference for functions which have, in addition,

3. $f(0)=0,$
4. $\lim_{x\rightarrow\infty}f(x)=\infty,$
5. $f$ is positively homogeneous, i.e. $f(\lambda x)=\lambda^\mu f(x)$ 
for all $\lambda\geq 0$ and for some $\mu>0$.
6. $f$ is convex.

Note that (5) implies (3) and (4). Power functions $x^r$ with $r>0$ clearly
satisfy the first five required and preferred conditions, and the sixth one
if $r\geq 1$. We will also consider the example of the logarithm (@ramsay_77), which does not satisfy preferred conditions (3),(5) and (6). An example of an $f$ that does not satisfy (4), (5) and (6) is the arctangent, bounded by $\pi/2$ as $x\rightarrow\infty$. The logarithm, the arctangent, and the power function with $r\leq 1$ are all concave.

For our purposes it is important that @groenen_deleeuw_mathar_C_95 show that at a local minimum
of fStress we have $d_k(x)>0$ for all $k$ for which $w_k\delta_k>0$, generalizing
the result for stress proved by @deleeuw_A_84f. Thus, under weak conditions
on the weights and dissimilarities, fStress is differentiable at
local minima.
@groenen_deleeuw_mathar_C_95 give no explicit algorithm to minimize fStress, but they do derive formulas for the first 
and second derivatives, which can of course be used in general-purpose optimization routines.

The important special case of fStress in which $f$ is a power function is called *rStress* (also known as *powerStress*). It is
\begin{equation}
\sigma_r(x):=\sum_{k=1}^K w_k(\delta_k^r-d_k^r(x))^2\label{eq-rdef}
\end{equation}
If $r=1$ we have Kruskal's stress, if $r=2$ we have *sstress* used by @takane_young_deleeuw_A_77. Using
\begin{equation}
\log(x)=\lim_{r\rightarrow 0}\frac{x^r-1}{r}\label{eq-loglim}
\end{equation}
shows that the logarithmic loss function used by @ramsay_77 is a limiting case of rStress. 

There have been various attempts to extend the smacof majorization (or MM) method for MDS (@deleeuw_C_77) to rStress. References and links to various unpublished reports are in @deleeuw_E_17r. The recent smacofx package (@rusch_deleeuw_mair_hornik_25) has R code for the rstressMin() function that implements the majorization method in @deleeuw_groenen_mair_E_16a.

Minimizing \eqref{eq-fdef1}
can be interpreted as approximating dissimilarities $\delta$ by distances $d(x)$ with a weighted least squares loss function, with $f$
incorporated in the weights. This can be seen most clearly if the fit is good, i.e. when
$\delta$ and $d(x)$ are close. We can then approximate fStress from \eqref{eq-fdef1}
as in
\begin{equation}
\sigma_f(x)\approx\sum_{k=1}^K w_k\{f'(\delta_k)\}^2(\delta_k-d_k(x))^2.
\label{eq-fapprox}\end{equation}
By modifying the weights in this way (@groenen_vandevelden_16) minimizing \eqref{eq-fapprox} can be done by standard metric smacof. This will presumably give a good approximation to minimizing fStress if the fit is good.

## Non-metric fStress

In non-metric MDS we minimize fStress not only over $x$ but also
over $\delta$, where $\delta$ is constrained to be in some subset $\Delta$
of $\mathbb{R}^K$. 

The most common case is ordinal MDS, where $\Delta$
is defined by the inequalities $\delta_1\leq\cdots\leq\delta_k$.
In the ordinal case we redefine fStress as
\begin{equation}
\sigma_f(x,\hat d):=\sum_{k=1}^Kw_k(\hat d_k-f(d_k(x)))^2,\label{eq-dhat}
\end{equation}
and we require that $\hat d\in\Delta\cap S$, where $S$ is the sphere
\begin{equation}
\sum_{k=1}^Kw_k\hat d_k^2=1.\label{eq-normal}
\end{equation}
In the MDS literature the $\hat d_k$ are called the *disparities* or 
the *pseudo-distances*. Note that it does not make a difference if
we require the disparities $\hat d$ to be monotone with the dissimilarities
$\delta$ or with the transformed dissimilarities $f(\delta)$.

It is important to observe that for rStress we have homogeneity





\sectionbreak

# Algorithm

The technique proposed in this paper to minimize non-metric fStress 
is in the Alternating Least Squares (ALS) family. We start with an initial estimate
$x^{(0)}$. We then minimize $\sigma_f$ over $\delta$ in $\Delta\cap\Sigma$ for 
the current $\smash{d(x^{(0)})}$. The minimizer $\smash{\delta^{(0)}}$ is then used to minimize
fStress over $x$ for the current disparities, yielding $x^{(1)}$. These two steps are
alternated until $x$ and $\delta$ do not change any more. Starting
in iteration $\nu=0$ we compute
\begin{subequations}
\begin{align}
\delta^{(\nu)}&=\mathop{\text{argmin}}_{\delta\in\Delta\cap\Sigma}\sigma_f(x^{(\nu)},\delta),\label{eq-als1}\\
x^{(\nu+1)}&=\mathop{\text{argmin}}_x\sigma_f(x,\delta^{(\nu)})\label{eq-als2}.
\end{align}
\end{subequations}
We then increase $\nu$ by one and go into the next iteration. 
And so on, until convergence.

## Normalized Cone Regression

In metric MDS there is no need for the first step \eqref{eq-als1}, because $\delta^{(\nu)}$ is equal to
the normalized dissimilarities $\delta$ for all $\nu$.
In the non-metric case step \eqref{eq-als1} is needed, but compared to \eqref{eq-als2} it is comparatively easy. We have to compute the least squares
projection of $f(d(x))$ on the cone $\Delta$ and then normalize this projection
to give it length one. If $\Delta$ is the cone of monotone vectors, as in
ordinal MDS, the cone projection is *monotone regression*. The fact that
projection on the intersection of $\Delta$ and $\Sigma$ is the same thing as
normalizing the projection on $\Delta$ is due to @deleeuw_U_75a 
and more generally (and more rigorously) to @bauschke_bui_wang_18.
Since this step is the same as in standard MDS algorithms such as
smacof (@deleeuw_mair_A_09c, @mair_groenen_deleeuw_A_22) we do not go
into details here, and just refer to the literature.

## Majorization

The second step \eqref{eq-als2}, minimizing over $x$ for fixed current $\delta$, is more complicated. There is
no analytic solution, unlike the first step, and minimization requires an iterative process of its own. Thus, except in some very special cases, the second step requires an infinite number of "inner" iterations. Since implementing an infinite number of iterations is impossible
we have to truncate the inner iteration sequence at some point. In our software we deviate from the strict ALS framework by not minimizing fStress over $x$ for fixed $\delta$,
but by taking a single "inner" iteration and merely decrease fStress. If this is
done judiciously we still obtain a convergent sequence of updates. This is similar to the strategy in other
MDS algorithms such as smacof (@deleeuw_C_77) and alscal (@takane_young_deleeuw_A_77).

In smacof the inner iteration step is a majorization step, by now more commonly known as an MM step.
Briefly, we find a function $\kappa_f(x;y)$ such that 

* $\kappa_f(x,y)\leq\sigma_f(x)$ for all $x$ and $y$ in $\Xi$, and 
* $\kappa_f(x;y)=\sigma_f(x)$ if and only if $x=y$.

An inner iteration is of the form
\begin{equation}
x^{(\mu+1)}=\mathop{\text{argmin}}_{x\in\Xi}\kappa(x,x^{(\mu)}).\label{eq-inner}
\end{equation}
If $x^{(\mu+1)}=x^{(\mu)}$ we declare convergence and stop.
From \eqref{eq-inner},
\begin{equation}
\sigma_f(x^{(\mu+1)})\leq\kappa_f(x^{(\mu+1)},x^{(\mu)})<\kappa_f(x^{(\mu)},x^{(\mu)})=\sigma_f(x^{(\mu)}).
\label{eq-sandwich}
\end{equation}
Thus an inner majorization step upgrading $x$ for given $\delta$ decreases
fStress (unless we stop). Remember that in ALS the inner iterative process (indexed by $\mu$) goes on within step 2 of the "outer" update (indexed by $\nu$). 

In @deleeuw_groenen_mair_E_16a a majorization method was proposed to minimize rStress.
It majorizes $d^r$ as a function of $x$, and 
for this it uses the specific properties of the power function. It consequently cannot be
used for fStress. The technique is incorporated in the smacofx R package, and 
numerical experience indicates it works, but convergence can be painfully slow.
Because of the generality of the function $f$ (any differentiable increasing function)
it seems impossible to develop an majorization method for fStress along the same lines as
the one for rStress.
Consequently we go a somewhat different route in this paper. We do not
only deviate from the strict ALS procedure by only partially minimizing
loss over $x$ for fixed $d$, we also deviate from the strict majorization
approach by majorizing an approximation of fStress. 

In deriving the approximation and majorization we surpress the dependency of fStress
on $\delta$, because in the second ALS substep $\delta$ is just a vector of constants.
We first approximate $f(d(x))$ near $d(y)$ with 
\begin{equation}
f(d_k(x))\approx f(d_k(y))+f;(d_k(y))(d_k(x)-d_k(y)).\label{eq-taylor}
\end{equation}
Define
\begin{equation}
\omega_f(x;y):=\sum_{k=1}^K w_k(f(\delta_k)-f(d_k(y))-f'(d_k(y))(d_k(x)-d_k(y)))^2,\label{eq-defeta}
\end{equation}
Note that in \eqref{eq-taylor} the derivative is with respect to $d$, not with respect to $x$ or $y$. This
is a major difference with the majorizations in @deleeuw_groenen_mair_E_16a.

Note that $\omega_f(x;x)=\sigma_f(x)$ for all $x$ and if $f$ is the identity then 
$\omega_f(x;y)=\sigma_f(x)$ for all $x$ and $y$. If $f$ is linear with intercept 
$\beta$ and slope $\alpha$ then 
\begin{equation}
\omega_f(x,y)=a^2\sum_{k=1}^Kw_k(\delta_k-d_k(x))^2=a^2\sigma_r(x),\label{eq-linear}
\end{equation}
with $r=1$.

We now give an alternative and more convenient expression for $\omega_f(x,y)$ from \eqref{eq-defeta}. Define
\begin{subequations}
\begin{align}
\tilde w_k(y)&:=w_k\{f'(d_k(y))\}^2,\label{eq-tildew}\\
\tilde\delta_k(y)&:=\frac{f(\delta_k)-f(d_k(y))}{f'(d_k(y))}+d_k(y).\label{eq-tilded}
\end{align}
\end{subequations}
Then
\begin{equation}
\omega_f(x;y)=\sum_{k=1}^K\tilde w_k(y)(\tilde\delta_k(y)-d_k(x))^2.\label{eq-omegas}
\end{equation}
We see from \eqref{eq-tildew} that $\tilde w_k(y)$ is always non-negative. If $f$ is convex
then $\tilde\delta_k(y)\geq\delta_k\geq 0$. If $f$ is concave then 
$\tilde\delta_k(y)\leq\delta_k$ and $\tilde\delta_k(y)$ can be negative. If
$f(\delta_k)=f(d_k(y))+f'(d_k(y))(\delta_k-d_k(y))+o(|\delta_k-d_k(y)|)$
then $\tilde\delta_k(y)=\delta_k+o(|\delta_k-d_k(y)|)$. Also
\begin{equation}
\tilde\delta_k(y)=\left\{\frac{f'(\xi)}{f'(d_k(y))}\right\}\delta_k+\left\{1-\frac{f'(\xi)}{f'(d_k(y))}\right\}d_k(y)
\end{equation}
for some $\xi$ in  the interval between $\delta_k$ and $d_k(y)$. 


For rStress equations \eqref{eq-tildew} and \eqref{eq-tilded} become
\begin{subequations}
\begin{align}
\tilde w_k(y)&:=r^2w_kd_k^{2(r-1)}(y),\label{eq-tildewr}\\
\tilde\delta_k(y)&:=\frac{\delta_k^r+(r-1)d_k^r(y)}{rd_k^{r-1}(y)}.\label{eq-tildedr}
\end{align}
\end{subequations}
By convexity $\tilde\delta_k(y)$ is non-negative if $r\geq 1$. For $0<r<1$ we have 
$\tilde\delta_k(y)>0$ if and only if $\smash{d_k(y)<(1-r)^r\delta_k}$. If $\delta_k>d_k(y)$
then $\tilde\delta_k(y)>d_k(y)\geq 0$.

By writing $\omega_f(x,y)$ as in \eqref{eq-omegas} we are on familiar majorization terrain.
If $\delta_k(y)>0$ we use Cauchy-Schwartz, as in standard smacof,
\begin{subequations}
\begin{equation}
d_k(x)\geq\frac{1}{d_k(y)}\text{tr}\ x'A_ky,\label{eq-cs}
\end{equation}
and if $\delta_k(y)<0$ we use the arithmetic mean/geometric mean (AM/GM) inequality
in the form
\begin{equation}
d_k(x)\leq\frac12\frac{1}{d_k(y)}\{y'A_ky+x'A_kx\},\label{eq-amgm}
\end{equation}
\end{subequations}
Use of the AM/GM inequality for majorizing terms with negative dissimilarities 
was pioneered by @heiser_91.

Define the matrices
\begin{subequations}
\begin{align}
V(y)&:=\sum_{k=1}^K\tilde w_k(y)A_k,\\
B(y)&:=\sum_{\tilde\delta_k(y)>0}\tilde w_k\frac{\tilde\delta_k(y)}{d_k(y)}A_k,\\
H(y)&:=\sum_{\tilde\delta_k(y)<0}\tilde w_k\frac{|\tilde\delta_k(y)|}{d_k(y)}A_k.
\end{align}
\end{subequations}
Note that all three matrices (more precisely matrix-valued functions) are positive semi-definite.
Now, using these matrices and the inequalities \eqref{eq-cs} and \eqref{eq-amgm},
\begin{equation}
\omega_f(x,y)\leq\kappa(x,y):=\gamma(y)+x'(V(y)+H(y))x-2x'B(y)y,
\label{eq-matmajx}
\end{equation}
with
\begin{equation}
\gamma(y):=\sum_{k=1}^K\tilde w_k(y)\tilde\delta_k^2(y)+y'H(y)y
\end{equation}
which does not depend on $x$.



We now update using
$$
x^{(\mu+1)}=\mathop{\text{argmin}}_x\kappa_f(x;x^{(\mu)})=(V(x^{(\mu)})+H(x^{(\mu)}))^+B(x^{(\mu)})x^{(\mu)},
$$
which results in
$$
\sigma_f(x^{(\mu+1)})\approx\omega_f(x^{(\mu+1)};x^{(\mu)})\leq\kappa_f(x^{(\mu+1)};x^{(\mu)})<\kappa_f(x^{(\mu)};x^{(\mu)})=\sigma_f(x^{(\mu)}).
$$
Unlike in smacof there is no guarantee that $\sigma_f$ decreases from one iteration to the other. Monotone
convergence of loss function values will only happen if the approximation is good, i.e. if
$d(x^{(\mu+1)})$ is close to $d_k(x^{(\mu)})$. This requires closer monitoring of convergence than 
in smacof. Also note that if $\tilde\delta_k(y)<0$ for all $k$ then $B(y)=0$, which means that the update 
from $y$ is equal to zero. Again, this eventuality must be monitored.

\sectionbreak

# Software


The R function smacofSSRStressR() and the C function smacofSSRStressC() in the files smacofSSFStressR.R and smacofSSFStressC.c handle both metric and ordinal MDS minimizing fStress, in principle for an arbitrary $f$ . The C code is compiled into a shared library smacofSSFStressC.so, which is loaded by the R front-end smacofSSFStressC.R

Both programs have input
```{r eval = FALSE}
function(theData,
         ndim = 2,
         xinit = NULL,
         ties = 1,
         itmax = 10000,
         usef = 1,
         eps = 1e-6,
         what = 0,
         rpow = 1,
         digits = 8,
         width = 10,
         verbose = TRUE,
         ordinal = FALSE)
```

Both functions output the same list
```{r eval = FALSE}
result <- list(
    delt = delt,
    dhat = dhat,
    dist = dnew,
    xmat = xnew,
    wmat = wght,
    loss = snew,
    ndim = ndim,
    init = xinit,
    itel = itel,
    nobj = nobj,
    iind = iind,
    jind = jind,
    ordi = ordinal,
    ties = ties,
    what = what,
    rpow = rpow,
    labl = labl, 
    usef = usef,
    date = date()
  )
```
The input parameters are copied in the output so that the result has all the
relevant information to reconstruct the run, and can pass it, for example, to
the plot routines.

## Data

For smacofSSRStressR.R  and smacofSSRStressC.R the data and weights are in the MDS data format defined in @deleeuw_E_25b. This is a list with two scalars nobj and ndat, the number of objects and the number of data points (object pairs), and five vectors iind, jind, delta, blocks, weights  of length ndat. Vectors iind and jind have integers between 1 and n with iind > jind that code for which pairs of objects we have dissimilarity information. The dissimilarities are in delta, the weights in weights. Dissimilarities are sorted in increasing order. The blocks vector codes the tie-blocks in the dissimilarities.

## Initial Estimate

There are two possibilities. If xinit is not NULL then it is used as the 
initial estimate of the configuration. This is useful, for example,
if we start the iterations for one $f$ with the solution for another.
If xinit is NULL then the function smacofSSRStressInit() computes an initial estimate using classical MDS (also known as Torgerson-Gower MDS) on the dissimilarities.

First we make sure the weights add up to one. If there are
missing dissimilarities they are first imputed by setting them equal to the 
weighted mean of the non-missing dissimilarities. Then we double center the squared $\delta$ and compute the classical scaling solution using
the eigs_sym() function from the RSpectra package (@qiu_mei_24).

It is well-known that if the fit is half-way decent then the distances in the Torgerson-Gower solution underestimate the dissimilarities. It makes sense,
consequently, to multiply the initial configuration by a constant to get a better fit (@malone_tarazaga_trosset_02). We use the fStress approximation \eqref{eq-fapprox} and minimize
$$
\sigma(\lambda)=\sum_{k=1}^Kw_k\{f'(\delta)\}^2(\delta_k-d_k(\lambda x))^2
$$
over $\lambda$.


## Functions

The R and C versions come with the following hard-coded functions. The
*what* (zero to four) selects the function, the *rpow* parameter
gives the $r$ parameter value. 


```{r kabf, echo = FALSE, eval = TRUE}

x <- matrix(
  c(
    "x^r",
    "rx^{r - 1}",
    "y^{1 / r}",
    "log(rx + 1)",
    "r/(rx+1)",
    "(exp(y) - 1) / r",
    "exp(rx) - 1",
    "r exp(rx)",
    "log(y + 1)/r",
    "rlog(x)",
    "r/x",
    "exp(y/r)",
    "atan(rx)",
    "r / (1 + (rx)^2)",
    "tan(y)/r"
  ),
  5,
  3,
  byrow = TRUE
)
rn <- c("power", "shifted log", "shifted exponent", "log", "atan")
row.names(x) <- rn
cn <- c("f(x)", "f'(x)", "f^{-1}(y)")
colnames(x) <- cn
kable(
  x,
  format = "simple",
  align = NULL,
  row.names = TRUE,
  caption = "f examples"
)
```

The selection code is in the
C function smacofSSFStressFlist() and in the R function
smcofSSFStressSelect(). It is easy to add 
functions to both versions, but adding them to the 
C version requires recompiling of the shared library.

## Arguments

The arguments of smacofSSFStressR() and smacofSSFStressR() are
```{r theargsf, echo = FALSE, eval = FALSE}
args(fStress)
```

Thus fStress() needs an initial configuration. Parameters $f$ and $g$ are functions that compute the transformation $f$ and its derivative. The file
fStress.R includes flog, glog, fexp, gexp, and fPower, gPower. Parameter
$r$ is included in case $f$ and $g$ need additional arguments (as in the case of 
fPower and gPower). Since all code is in R, it is easy to add additional 
transformation functions if needed. The output of fStress() is 
```{r outputf, eval = FALSE}
list(
    delt = delta, 
    dhat = delf, 
    wmat = wght,
    conf = xnew,
    dist = dnew,
    loss = snew,
    itel = itel
  )
```

The arguments of smacofSSRStress() are
```{r theargsr, echo = FALSE, eval = FALSE}
args(smacofSSRStress)
```


## Value

The C code follows the conventions of the smacof versions in @deleeuw_E_25b. 
There is only a single parameter $rpow$ for the transformation. The output
of smacofSSRStress() is
```{r output, eval = FALSE}
list(
    delt = theData$delta,
    dhat = h$dhat,
    dist = h$edis,
    xmat = matrix(h$xnew, nobj, ndim),
    wmat = h$wght,
    loss = h$snew,
    ndim = ndim,
    init = xinit,
    itel = h$itel,
    nobj = nobj,
    iind = h$iind,
    jind = h$jind,
    weighted = weighted,
    ordi = ordinal,
    ties = h$ties
  )
```



## Plots

The four standard plots that come with the recent smacof programs () are

* smacofConfigurationPlot()
* smacofShepardPlot()
* smacofDistDhatPlot()
* smacofResidualPlot()

The output list of smacofSSFStress() can be fed immediately into these four plot routines. They only have side effects (i.e. make a plot) and return NULL.

###  smacofConfigurationPlot()

It is easy to explain what smacofConfigurationPlot() does. It plots a two-dimensional
projection of the configuration. Its arguments are self-explanatory.
```{r eval = FALSE}
function(h, labels = NULL, dim1 = 1, 
         dim2 = 2, pch = 16, col = "RED", cex = 1) 
```

### smacofShepardPlot()

A Shepard plot (@deleeuw_mair_C_15) traditionally plots observed dissimilarities on the horizontal axis, and both disparities and distances on the vertical axis.
The $(\delta_k^0,\delta_k)$ points are connected by a straight line through the origin in the metric case and by an increasing step function in the ordinal case. The $(delta,dist)$ points
are optionally connected with vertical "fitlines" to the corresponding $(delta,dhat)$ points.
```{r eval = FALSE}
function(h, main = "ShepardPlot", fitlines = TRUE, colline = "RED", 
    colpoint = "BLUE", type = 0, lwd = 2, cex = 1, pch = 16) 
```

In the fStress solution it makes sense to put the final $\delta$ and 
$f(d(x))$ on the vertical axis, but there is some ambiguity on what
we use on the horizontal axis. As explained in the introduction in metric MDS we can have $\delta\propto f(\delta^0)$, with $\delta^0$ observed dissimilarities.
In that case we put $f(\delta^0)$ on the horizontal axis, and
the the points $f(\delta_k^0),\delta_k$ are on a straight line.

In the numerical case do we analyze $\delta=f(\delta^0)$ or $\delta=\delta^0$.
In the ordinal case: it does not make a difference, except for the plots

In the fStress context there is an extra arguments "type".
Remember that the list h has both a "what" and an "rpow" element.
The "type" argument is either one or zero. If it is zero (the default) then
we plot $f(\delta_k)$ on the horizontal axis and $\hat d_k$ and $f(d_k(x))$ on the vertical axis. If "type" is equal to one then we plot $\delta_k$ on the horizontal axis and $f^{-1}(\hat d_k)$ and $d_k(x)$ on the vertical axis.

### smacofDistDhatPlot()

smacofDistDhatPlot() has arguments
```{r eval = FALSE}
function(h, fitlines = TRUE, colline = "RED", colpoint = "BLUE", 
    main = "Dist-Dhat Plot", type = 0, cex = 1, lwd = 2, pch = 16) 
```
The "type" parameter works in the same way as in the Shepard plot.

### smacofResidualPlot()

smacofResidualPlot() is more complicated. It has arguments
```{r eval = FALSE}
function (h, main = "ResidualPlot", probs = seq(0, 1, 0.25), 
    whatres = 1, whatplot = 1, type = 0, qlines = TRUE, colpoints = "RED", 
    collines = "BLUE", lwdpoints = 2, lwdlines = 2, cex = 1, 
    pch = 16) 
```

As in the previous plots *type* computes residuals as 

*whatplot* can have the values *ecdf* (default) or *density*. Computation is done with the ecdf() and density() functions from the stats package.

*whatres* chooses to plot the residual (default), its absolute value, or its square

\sectionbreak

# Example

## Functions

We analyse the colour data from @ekman_54 with seven different functions $f$, using both 
smacofSSFStressR() and using smacofSSFStressC(), both metric and ordinal.

```{r rfstressruns, echo = FALSE, cache = FALSE, eval = FALSE}
hl <- smacofSSFStressR(ekmanData, what = 1, verbose = FALSE)
hp1 <- smacofSSFStressR(ekmanData, rpow = .1, verbose = FALSE)
hp5 <- smacofSSFStressR(ekmanData, rpow = .5, verbose = FALSE)
h1 <- smacofSSFStressR(ekmanData, rpow = 1, verbose = FALSE)
h2 <- smacofSSFStressR(ekmanData, rpow = 2, verbose = FALSE)
h5 <- smacofSSFStressR(ekmanData, rpow = 5, verbose = FALSE)
he <- smacofSSFStressR(ekmanData, what = 2, verbose = FALSE)
```
```{r kableR, echo = FALSE}
rloss <- c(0.00418, 0.00026, 0.00467, 0.01160,0.01824,0.00803,0.04511)
riter <- c(154, 804, 227, 35,49,793,36)
x <- matrix(c(riter, rloss), 7, 2)
rnames <- c("shifted logarithm", 
       "power .1",
       "power .5",
       "power 1",
       "power 2", 
       "power 5",
       "shifted exponent"
)
row.names(x) <- rnames
cnames <- c("itel", "fStress")
colnames(x) <- cnames
kable(x, format = "simple", align = "r", row.names = TRUE, caption = "smacofSSFStressC() results Ekman Data")
```

```{r kableRO, echo = FALSE}
rlosso <- c(0.0002466493, 6.567816e-06, 0.0001651214, 0.0005337258, 0.0009014502, 0.001100227, 0.0007894837)
ritero <-  c(336, 980, 213, 141, 231, 2625, 159)
x <- matrix(c(ritero, rlosso), 7, 2)
rnames <- c(
       "shifted logarithm",
       "power .1",
       "power .5",
       "power 1",
       "power 2", 
       "power 5",
       "shifted exponential"
)
row.names(x) <- rnames
cnames <- c("itel", "fStress")
colnames(x) <- cnames
kable(x, format = "simple", align = "r", row.names = TRUE, caption = "
      smacofSSFStressC() results Ekman Data, ordinal MDS")
```

## Timings

The package microbenchmark (@mersmann_24) can be used to compare the
R and C versions. We report the median time for 100 runs of each
anaysis. In the metric case the C version is between 10 and
30 times faster than the R version, with the smallest difference for 
the stress ($r=1$) and sstress ($r=2$) solutions. In the non-metric
case the C version is consistently between 40 and 50 times faster.
In packages such as smacof and smacofx the majorization step is 
dome in R and the monotone regression step is done in C. In that
case the speedup of our all-C version will be somewhere between
the metric and non-metric speedsups in the tables.


```{r kablemicrometric, echo = FALSE}
rtimes <- c(14.992, 80.865, 22.245, 3.641, 4.947, 79.643, 3.707)
ctimes <-  c(0.575, 2.984, 1.040, 0.366, 0.431, 2.900, 0.340)
ratio <- c(26, 27, 21, 10, 11, 27, 11)
x <- matrix(c(rtimes, ctimes, ratio), 7, 3)
rnames <- c(
       "shfted logarithm",
       "power .1",
       "power .5",
       "power 1",
       "power 2", 
       "power 5",
       "shifted exponential"
)
row.names(x) <- rnames
cnames <- c("R times", "C times", "R/C")
colnames(x) <- cnames
kable(x, format = "simple", align = "r", row.names = TRUE, caption = "
      R and C timings Ekman Data in milliseconds, metric MDS")
```
```{r eval = FALSE, echo = FALSE}
Unit: microseconds
                                                     expr       min          lq        mean
   smacofSSFStressR(ekmanData, what = 1, verbose = FALSE) 23038.638  23619.8540  28644.5889
   smacofSSFStressC(ekmanData, what = 1, verbose = FALSE)   618.198    668.6690    738.0201
 smacofSSFStressR(ekmanData, rpow = 0.1, verbose = FALSE) 99130.046 100679.3745 111889.0709
 smacofSSFStressC(ekmanData, rpow = 0.1, verbose = FALSE)  2901.857   3071.4125   3152.8094
 smacofSSFStressR(ekmanData, rpow = 0.5, verbose = FALSE) 26666.523  27385.2325  31396.8271
 smacofSSFStressC(ekmanData, rpow = 0.5, verbose = FALSE)  1011.757   1058.1485   1090.9260
   smacofSSFStressR(ekmanData, rpow = 1, verbose = FALSE)  3676.839   3922.1010   5143.4389
   smacofSSFStressC(ekmanData, rpow = 1, verbose = FALSE)   340.464    361.7635    401.6942
   smacofSSFStressR(ekmanData, rpow = 2, verbose = FALSE)  5583.913   5854.9640   6903.7317
   smacofSSFStressC(ekmanData, rpow = 2, verbose = FALSE)   407.048    438.7000    481.1440
   smacofSSFStressR(ekmanData, rpow = 5, verbose = FALSE) 92626.995  95049.3980 100948.4161
   smacofSSFStressC(ekmanData, rpow = 5, verbose = FALSE)  2747.779   2863.9730   2987.3613
   smacofSSFStressR(ekmanData, what = 2, verbose = FALSE)  3689.795   3823.3525   4600.4087
   smacofSSFStressC(ekmanData, what = 2, verbose = FALSE)   306.270    335.2160    406.1378
      median          uq        max neval    cld
  24030.0180  27291.9780  86524.760   100 a     
    690.3580    734.2280   4413.978   100  bc   
 101870.7320 105006.2890 220077.627   100    d  
   3115.0980   3169.8125   6647.986   100   c e 
  27918.6220  31085.1340  90309.224   100 a     
   1073.4005   1118.4800   1236.847   100  bc   
   3966.5040   4026.5485  65376.304   100  b  e 
    381.6485    439.1715    510.163   100   c   
   5929.2355   6030.7515  68230.150   100     e 
    474.4315    523.5905    611.720   100   c   
  95857.4055  97766.0990 158889.965   100      f
   2919.5690   2962.3525   6655.858   100   c e 
   3882.6795   3933.7860  65264.415   100   c e 
    349.4020    405.1620   4051.948   100   c   
```
```{r kablemicrononmetric, echo = FALSE}
rtimes <- c(113.944, 430.655, 112.219, 58.316, 96.421, 1178.645, 64.072)
ctimes <-  c(2.819, 9.030, 2.245, 1.466, 2.366, 23.513, 1.500)
ratio <- c(40, 48, 50, 40, 41, 50, 43)
x <- matrix(c(rtimes, ctimes, ratio), 7, 3)
rnames <- c(
       "shifted logarithm",
       "power .1",
       "power .5",
       "power 1",
       "power 2", 
       "power 5",
       "shifted exponential"
)
row.names(x) <- rnames
cnames <- c("R times", "C times", "R/C")
colnames(x) <- cnames
kable(x, format = "simple", align = "r", row.names = TRUE, caption = "
      R and C timings Ekman Data in milliseconds, nonmetric MDS")
```
```{r eval = FALSE, echo = FALSE}
Unit: milliseconds
                                                                     expr         min          lq
   smacofSSFStressR(ekmanData, what = 1, verbose = FALSE, ordinal = TRUE)  106.306973  110.755452
   smacofSSFStressC(ekmanData, what = 1, verbose = FALSE, ordinal = TRUE)    2.542246    2.768279
 smacofSSFStressR(ekmanData, rpow = 0.1, verbose = FALSE, ordinal = TRUE)  402.058095  414.841136
 smacofSSFStressC(ekmanData, rpow = 0.1, verbose = FALSE, ordinal = TRUE)    8.338334    8.780068
 smacofSSFStressR(ekmanData, rpow = 0.5, verbose = FALSE, ordinal = TRUE)  109.666185  114.417306
 smacofSSFStressC(ekmanData, rpow = 0.5, verbose = FALSE, ordinal = TRUE)    1.954347    2.092128
   smacofSSFStressR(ekmanData, rpow = 1, verbose = FALSE, ordinal = TRUE)   55.808298   57.330997
   smacofSSFStressC(ekmanData, rpow = 1, verbose = FALSE, ordinal = TRUE)    1.311385    1.401995
   smacofSSFStressR(ekmanData, rpow = 2, verbose = FALSE, ordinal = TRUE)   89.000996   93.787131
   smacofSSFStressC(ekmanData, rpow = 2, verbose = FALSE, ordinal = TRUE)    2.181077    2.311272
   smacofSSFStressR(ekmanData, rpow = 5, verbose = FALSE, ordinal = TRUE) 1118.418623 1155.259296
   smacofSSFStressC(ekmanData, rpow = 5, verbose = FALSE, ordinal = TRUE)   22.259269   22.983288
   smacofSSFStressR(ekmanData, what = 2, verbose = FALSE, ordinal = TRUE)   59.894727   62.936353
   smacofSSFStressC(ekmanData, what = 2, verbose = FALSE, ordinal = TRUE)    1.383791    1.457017
        mean      median          uq         max neval       cld
  113.852865  111.933280  113.670327  176.549526   100 a        
    2.911093    2.826417    2.886216    6.084523   100  b       
  423.109740  420.029461  424.704014  487.229322   100   c      
    8.936843    8.955302    9.134964    9.351895   100    d     
  117.196493  115.901117  117.756203  179.792503   100 a        
    2.184620    2.150245    2.207522    5.484734   100  b       
   58.943635   57.998190   58.850642  122.780691   100     e    
    1.450922    1.456792    1.499903    1.597770   100  b       
   95.140413   95.257411   96.381037   99.670098   100      f   
    2.402289    2.382243    2.436856    5.536845   100  b       
 1176.369676 1172.620849 1186.844917 1265.874795   100       g  
   23.475012   23.533077   23.939449   24.703853   100        h 
   63.974907   63.568348   64.972679   68.236505   100         i
    1.497407    1.492666    1.539263    1.626183   100  b       
```

## Plots

\sectionbreak
```{r chp1c, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Metric, r = .1", main ="", eval = FALSE}
par(pty="s")
smacofConfigurationPlot(chp1, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r chp1s, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Metric, r = .1", main = "", eval = FALSE}
smacofShepardPlot(chp1, rpow = .1, cex = 2)
```
\sectionbreak
```{r chp5c, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Metric, r = .5", main ="", eval = FALSE}
par(pty="s")
smacofConfigurationPlot(chp5, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r chp5s, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Metric, r = .5", main = "", eval = FALSE}
smacofShepardPlot(chp5, rpow = .5, cex = 2)
```
\sectionbreak
```{r ch1c, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Metric, r = 1", main ="", eval = FALSE}
par(pty="s")
smacofConfigurationPlot(ch1, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r ch1s, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Metric, r = 1", main = "", eval = FALSE}
smacofShepardPlot(ch1, rpow = 1, cex = 2)
```
\sectionbreak
```{r ch2c, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Metric, r = 2", main ="", eval = FALSE}
par(pty="s")
smacofConfigurationPlot(ch2, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r ch2s, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Metric, r = 2", main = "", eval = FALSE}
smacofShepardPlot(ch2, rpow = 2, cex = 2)
```
\sectionbreak
```{r ch5c, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Metric, r = 5", main ="", eval = FALSE}
par(pty = "s")
smacofConfigurationPlot(ch5, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r ch5s, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Metric, r = 5", main = "", eval = FALSE}
smacofShepardPlot(ch5, rpow = 5, cex = 2)
```

\sectionbreak
```{r chp1oc, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Non-metric, r = .1", main ="", eval = FALSE}
par(pty="s")
smacofConfigurationPlot(chp1o, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r chp1os, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Non-metric, r = .1", main = "", eval = FALSE}
smacofShepardPlot(chp1o, rpow = 2, cex = 2)
```
\sectionbreak
```{r chp5oc, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Non-metric, r = .5", main ="", eval = FALSE}
par(pty="s")
smacofConfigurationPlot(chp5o, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r chp5os, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Non-metric, r = .5", main = "", eval = FALSE}
smacofShepardPlot(chp5o, rpow = .5, cex = 2)
```
\sectionbreak
```{r ch1oc, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Non-metric, r = 1", main ="", eval = FALSE}
par(pty = "s")
smacofConfigurationPlot(ch1o, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r ch1os, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Non-metric, r = 1", main = "", eval = FALSE}
smacofShepardPlot(ch1o, rpow = 1, cex = 2)
```
\sectionbreak
```{r ch2oc, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Non-metric, r = 2", main ="", eval = FALSE}
par(pty = "s")
smacofConfigurationPlot(ch2o, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r ch2os, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Non-metric, r = 2", main = "", eval = FALSE}
smacofShepardPlot(ch2o, rpow = 2, cex = 2)
```
\sectionbreak
```{r ch5oc, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Non-metric, r = 5", main ="", eval = FALSE}
par(pty = "s")
smacofConfigurationPlot(ch5o, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r ch5os, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Non-metric, r = 5", main = "", eval = FALSE}
smacofShepardPlot(ch5o, rpow = 5, cex = 2)
```


\sectionbreak

# References