---
title: Minimizing fStress and rStress by Majorizing Gauss-Newton
author: 
    - name: Jan de Leeuw
      orcid: 0000-0003-1420-1797
      email: jan@deleeuwpdx.net
      affiliation: 
        - name: University of California Los Angeles
          city: Los Angeles
          state: CA
          url: www.ucla.edu
      license: "CC0"
date: last-modified
date-format: long
bibliography: [mypubs.bib, total.bib]
number-sections: true
pdf-engine: lualatex
keep-tex: true
format:
   pdf:
    fontsize: 12pt
    include-in-header: 
     - preamble.tex
    keep-tex: true
    link-citations: true
    documentclass: scrartcl
    number-sections: true
   html:
    fontsize: 12pt
    include-in-header: 
     - preamble.css
    keep-md: true
    number-sections: true
toc: true
toc-depth: 3
editor: source
papersize: letter
graphics: true
link-citations: true
mainfont: Times New Roman
abstract: TBD
---

```{r loadpackages, echo = FALSE}
suppressPackageStartupMessages(library(knitr, quietly = TRUE))
suppressPackageStartupMessages(library(tinytex, quietly = TRUE))
suppressPackageStartupMessages(library(microbenchmark, quietly = TRUE))
```

```{r load sources, echo = FALSE}
source("smacofSSRStress.R")
source("smacofPlots.R")
source("fStress.R")
source("ekmanData.R")
```

\sectionbreak 

**Note:** This is a working manuscript which will be expanded/updated
frequently. All suggestions for improvement are welcome. All Rmd, tex,
html, pdf, R, and C files are in the public domain. Attribution will be
appreciated, but is not required. The files can be found at
<https://github.com/deleeuw/rStress> 

\sectionbreak

# Loss Functions

The Multidimensional Scaling (MDS) loss function fStress is defined as
\begin{equation}
\sigma_f(x):=\frac12\sum_{k=1}^K w_k(f(\delta_k)-f(d_k(x)))^2,\label{eq-fdef}
\end{equation}
with $f$ increasing and differentiable in the open interval $(0,+\infty)$. 
In \eqref{eq-fdef} the $w_k$ are positive *weights*, the $\delta_k$ are 
known *dissimilarities*. The vector $x$ has the coordinates of $n$
points in $\mathbb{R}^p$, and the $d_k(x)$ are Euclidean *distances* between pairs 
of these points. For each $k$ there is a pair of indices $(i,j)$ with $1\leq i<j\leq n$
and a matrix $A_k$ which is the direct sum of $p$ copies of $(e_i-e_j)(e_i-e_j)'$,
where the $e_i$ are unit vectors (columns of the identity matrix of order $n$).
Thus $\smash{d_k(x)=\sqrt{x'A_kx}}$. Metric least squares MDS minimizes fStress over $x$.

fStress was first introduced and studied in @groenen_deleeuw_mathar_C_95. No explicit
algorithm to minimize loss \eqref(eq-fdef) was given, but the paper has formulas for the first 
and second derivatives (see also @deleeuw_E_17r), which can of course be used in general purpose optimization routines.

An important special case of fStress is rStress (also known as powerStress), which is
\begin{equation}
\sigma_r(x):=\frac12\sum_{k=1}^K w_k(\delta_k^r-d_k^r(x))^2\label{eq-rdef}
\end{equation}
Special cases of rStress are Kruskal's stress (@kruskal_64a, @kruskal_64b) with $r=1$, sstress by @takane_young_deleeuw_A_77 with $r=2$, and the limiting case logarithmic stress with $r\rightarrow 0$ by @ramsay_77. 
There have been various attempts to extend the majorization (or MM) method for MDS (@deleeuw_C_77) to rStress. References and links to various unpublished reports are in @deleeuw_E_17r. The recent smacofx package (@rusch_deleeuw_mair_hornik_25) has R code for the rstressMin() function that implements the majorization method in @deleeuw_groenen_mair_E_16a.

Minimizing of either fStress or rStress over $x$ is a metric MDS problem. The
$\delta_k$, and consequently the $f(\delta_k)$, are $K$ known numbers. In non-metric
MDS the loss function is minimized over both $x$ and $\delta$, where $\delta$
is constrained to be in a subset $\Delta$ of $\mathbb{R}^K$. For the ordinal 
version of non-metric MDS, for example, we require $\delta_1\leq\cdots\leq\delta_K$. 
Since $f$ is increasing, we can write fStress simply as
\begin{equation}
\sigma_f(x,\delta):=\frac12\sum_{k=1}^K w_k(\delta_k-f(d_k(x)))^2,\label{eq-nmfdef}
\end{equation}
where $\delta$ is no longer a vector of known dissimilarities, but any vector
monotone with the dissimilarities. These transformed or scaled dissimilarities
are often called *disparities*. Non-metric rStress is simply \eqref{eq-nmfdef}
with $\smash{f(d_k(x))=d_k^r(x)}$.

In order to exclude the trivial solution
with $x=0$ and $\delta=0$ in addition we impose the normalization constraint
\begin{equation}
\eta^2(\delta):=\frac12\sum_{k=1}^Kw_k\delta_k^2=1.\label{eq-normal}
\end{equation}
This formulation of non-metric MDS can be generalized to $\delta\in\Delta$,
where $\Delta$ is a convex cone in $\mathbb{R}^K$. This means we require
the disparities to be in the intersection of the cone $\Delta$ and the sphere $\Sigma$ defined
by \eqref{eq-normal}.

It is convenient to think of metric MDS as the special case in which the 
set $\Delta$ is the one-element set containing only the normalized dissimilarities
$\delta$. Alternatively, as in smacof, we can choose for $\Delta$ the ray of all vectors
that are non-negative multiples of the dissimilarities. In other words our metric
MDS treats the dissimilarities as measured on a ratio scale.


# Algorithm

The technique proposed in this paper to minimize non-metric fStress or rStress
is in the Alternating Least Squares (ALS) family. We start with an initial estimate
$x^{(0)}$. We then minimize $\sigma_f$ over $\delta$ in $\Delta\cap\Sigma$ for 
the current $\smash{d(x^{(0)})}$. The minimizer $\smash{\delta^{(0)}}$ is then used to minimize
fStress over $x$ for the current disparities, yielding $x^{(1)}$. These two steps are
alternated until $x$ and $\delta$ do not change any more. Starting
in iteration $\nu=0$ we compute
\begin{subequations}
\begin{align}
\delta^{(\nu)}&=\mathop{\text{argmin}}_{\delta\in\Delta\cap\Sigma}\sigma_f(x^{(\nu)},\delta),\label{eq-als1}\\
x^{(\nu+1)}&=\mathop{\text{argmin}}_x\sigma_f(x,\delta^{(\nu)})\label{eq-als2}.
\end{align}
\end{subequations}
We then increase $\nu$ by one and go into the next iteration. 
And so on, until convergence.


## Normalized Cone Regression

In metric MDS there is no need for the first step \eqref{eq-als1}, because $\delta^{(\nu)}$ is equal to
the normalized dissimilarities $\delta$ for all $\nu$.
In the non-metric case step \eqref{eq-als1} is needed, but compared to \eqref{eq-als2} it is comparatively easy. We have to compute the least squares
projection of $f(d(x))$ on the cone $\Delta$ and then normalize this projection
to give it length one. If $\Delta$ is the cone of monotone vectors, as in
ordinal MDS, the cone projection is *monotone regression*. The fact that
projection on the intersection of $\Delta$ and $\Sigma$ is the same thing as
normalizing the projection on $\Delta$ is due to @deleeuw_U_75a 
and more generally (and more rigorously) to @bauschke_bui_wang_18.
Since this step is the same as in standard MDS algorithms such as
smacof (@deleeuw_mair_A_09c, @mair_groenen_deleeuw_A_22) we do not go
into details here, and just refer to the literature.

## Majorization

The second step \eqref{eq-als2}, minimizing over $x$ for fixed current $\delta$, is more complicated. There is
no analytic solution, unlike the first step, and minimization requires an iterative process of its own. Thus, except in some very special cases, the second step requires an infinite number of "inner" iterations. Since implementing an infinite number of iterations is impossible
we have to truncate the inner iteration sequence at some point. In our software we deviate from the strict ALS framework by not minimizing fStress over $x$ for fixed $\delta$,
but by taking a single "inner" iteration and merely decrease fStress. If this is
done judiciously we still obtain a convergent sequence of updates. This is similar to the strategy in other
MDS algorithms such as smacof (@deleeuw_C_77) and alscal (@takane_young_deleeuw_A_77).

In smacof the inner iteration step is a majorization step, by now more commonly known as an MM step.
Briefly, we find a function $\kappa_f(x;y)$ such that 

* $\kappa_f(x,y)\leq\sigma_f(x)$ for all $x$ and $y$ in $\Xi$, and 
* $\kappa_f(x;y)=\sigma_f(x)$ if and only if $x=y$.

An inner iteration is of the form
\begin{equation}
x^{(\mu+1)}=\mathop{\text{argmin}}_{x\in\Xi}\kappa(x,x^{(\mu)}).\label{eq-inner}
\end{equation}
If $x^{(\mu+1)}=x^{(\mu)}$ we declare convergence and stop.
From \eqref{eq-inner},
\begin{equation}
\sigma_f(x^{(\mu+1)})\leq\kappa_f(x^{(\mu+1)},x^{(\mu)})<\kappa_f(x^{(\mu)},x^{(\mu)})=\sigma_f(x^{(\mu)}).
\label{eq-sandwich}
\end{equation}
Thus an inner majorization step upgrading $x$ for given $\delta$ decreases
fStress (unless we stop). Remember that in ALS the inner iterative process (indexed by $\mu$) goes on within step 2 of the "outer" update (indexed by $\nu$). 

In @deleeuw_groenen_mair_E_16a a majorization method was proposed to minimize rStress.
It majorizes $d^r$ as a function of $x$, and 
for this it uses the specific properties of the power function. It consequently cannot be
used for fStress. The technique is incorporated in the smacofx R package, and 
numerical experience indicates it works, but convergence can be painfully slow.
Because of the generality of the function $f$ (any differentiable increasing function)
it seems impossible to develop an majorization method for fStress along the same lines as
the one for rStress.
Consequently we go a somewhat different route in this paper. We do not
only deviate from the strict ALS procedure by only partially minimizing
loss over $x$ for fixed $d$, we also deviate from the strict majorization
approach by majorizing an approximation of fStress. 

In deriving the approximation and majorization we surpress the dependency of fStress
on $\delta$, because in the second ALS substep $\delta$ is just a vector of constants.
We first approximate $f(d(x))$ near $d(y)$ with 
\begin{equation}
f(d_k(x))\approx f(d_k(y))+\mathcal{D}f(d_k(y))(d_k(x)-d_k(y)).\label{eq-taylor}
\end{equation}
Define
\begin{equation}
\omega_f(x;y):=\sum_{k=1}^K w_k(f(\delta_k)-f(d_k(y))-\mathcal{D}f(d_k(y))(d_k(x)-d_k(y)))^2,\label{eq-defeta}
\end{equation}
Note that in \eqref{eq-taylor} the derivative is with respect to $d$, not with respect to $x$ or $y$. This
is a major difference with the majorizations in @deleeuw_groenen_mair_E_16a.

Note that $\omega_f(x;x)=\sigma_f(x)$ for all $x$ and if $f$ is the identity then 
$\omega_f(x;y)=\sigma_f(x)$ for all $x$ and $y$. If $f$ is linear with intercept 
$\beta$ and slope $\alpha$ then 
\begin{equation}
\omega_f(x,y)=a^2\sum_{k=1}^Kw_k(\delta_k-d_k(x))^2=a^2\sigma_r(x),\label{eq-linear}
\end{equation}
with $r=1$.

We now give an alternative and more convenient expression for $\omega_f(x,y)$. Define
\begin{subequations}
\begin{align}
\tilde w_k(y)&:=w_k\{\mathcal{D}f(d_k(y))\}^2,\label{eq-tildew}\\
\tilde\delta_k(y)&:=\frac{f(\delta_k)-f(d_k(y))}{\mathcal{D}f(d_k(y))}+d_k(y).\label{eq-tilded}
\end{align}
\end{subequations}
Then
\begin{equation}
\omega_f(x;y)=\sum_{k=1}^K\tilde w_k(y)(\tilde\delta_k(y)-d_k(x))^2.\label{eq-omegas}
\end{equation}
We see from \eqref{eq-tildew} that $\tilde w_k(y)$ is always non-negative and from \eqref{eq-tilded}
that $\tilde\delta_k(y)$ can be negative if $d_k(y)>\delta_k$. For convex $f$, however, we have
\begin{equation}
\tilde\delta_k(y)=\frac{f(\delta_k)-f(d_k(y))}{\mathcal{D}f(d_k(y))}+d_k(y)\geq d_k(x)\geq 0.\label{eq-convex}
\end{equation}

For rStress equations \eqref{eq-tildew} and \eqref{eq-tilded} become
\begin{subequations}
\begin{align}
\tilde w_k(y)&:=r^2w_kd_k^{2(r-1)}(y),\label{eq-tildewr}\\
\tilde\delta_k(y)&:=\frac{\delta_k^r+(r-1)d_k^r(y)}{rd_k^{r-1}(y)}.\label{eq-tildedr}
\end{align}
\end{subequations}
By convexity $\tilde\delta_k(y)$ is non-negative if $r\geq 1$. For $0<r<1$ we have 
$\tilde\delta_k(y)>0$ if and only if $\smash{d_k(y)<(1-r)^r\delta_k}$. If $\delta_k>d_k(y)$
then $\tilde\delta_k(y)>d_k(y)\geq 0$.

By writing $\omega_f(x,y)$ as in \eqref{eq-omegas} we are on familiar majorization terrain.
If $\delta_k(y)>0$ we use Cauchy-Schwartz, as in standard smacof,
\begin{subequations}
\begin{equation}
d_k(x)\geq\frac{1}{d_k(y)}\text{tr}\ x'A_ky,\label{eq-cs}
\end{equation}
and if $\delta_k(y)<0$ we use the arithmetic mean/geometric mean (AM/GM) inequality
in the form
\begin{equation}
d_k(x)\leq\frac12\frac{1}{d_k(y)}\{y'A_ky+x'A_kx\},\label{eq-amgm}
\end{equation}
\end{subequations}
Use of the AM/GM inequality for majorizing terms with negative dissimilarities 
was pioneered by @heiser_91.

Define the matrices
\begin{subequations}
\begin{align}
V(y)&:=\sum_{k=1}^K\tilde w_k(y)A_k,\\
B(y)&:=\sum_{\tilde\delta_k(y)>0}\tilde w_k\frac{\tilde\delta_k(y)}{d_k(y)}A_k,\\
H(y)&:=\sum_{\tilde\delta_k(y)<0}\tilde w_k\frac{\tilde\delta_k(y)}{d_k(y)}A_k.
\end{align}
\end{subequations}
Note that all three matrices (more preecisely matrix-valued functions) are positive semi-definite.
We can use these matrices for a majorization of $\omega_f(x;y)$ at $y$ is
\begin{equation}
\omega_f(x;y)=C+\frac12 x'(V(y)+H(y))x-x'B(y)x\leq\kappa_f(x;y),\label{eq-matmajx}
\end{equation}
with
\begin{equation}
\kappa_f(x;y):=C+\frac12 x'(V(y)+H(y))x-x'B(y)y,\label{eq-matmajy}
\end{equation}
where $C$ is a constant that depends on $y$ but not on $x$.

We now update using
$$
x^{(\mu+1)}=\mathop{\text{argmin}}_x\kappa_f(x;x^{(\mu)})=(V(x^{(\mu)})+H(x^{(\mu)}))^+B(x^{(\mu)})x^{(\mu)},
$$
which results in
$$
\sigma_f(x^{(\mu+1)})\approx\omega_f(x^{(\mu+1)};x^{(\mu)})\leq\kappa_f(x^{(\mu+1)};x^{(\mu)})<\kappa_f(x^{(\mu)};x^{(\mu)})=\sigma_f(x^{(\mu)}).
$$
Unlike in smacof there is no guarantee that $\sigma_f$ decreases from one iteration to the other. Monotone
convergence of loss function values will only happen if the approximation is good, i.e. if
$d(x^{(\mu+1)})$ is close to $d_k(x^{(\mu)})$. This requires closer monitoring of convergence than 
in smacof. Also note that if $\tilde\delta_k(y)<0$ for all $k$ then $B(y)=0$, which means that the update 
from $y$ is equal to zero. Again, this eventuality must be monitored.

# Software

There are two different major routines in the repository. The R function
fStress.R does metric MDS minimizing fStress with an arbitrary $f$. The C function
smacofSSRStress.c does either metric or ordinal MDS minimizing rStress for
an arabitrary $r$. The C code is compiled into a shared library smacofSSRStress.so, which is loaded by the R front-end smacofSSRStress.R
that contains smacofSSRStress().
The function fStress() is meant for exploration and research, while
smacofSSRStress() is more suitable for production. fStress() is more general, in
that it allows for more transformations, but smacofSSRStress() is more general
in that it allows for ordinal MDS.


## Data

For fStress.R the input is a matrix delta of dissimilarities. For smacofSSRStress.R the data and weights are in the MDS data format defined in @deleeuw_E_25b. This is a list with two scalars nobj and ndat, the number of objects and the number of data points (object pairs), and five vectors iind, jind, delta, blocks, weights  of length ndat. Vectors iind and jind have integers between 1 and n with iind > jind that code for which pairs of objects we have dissimilarity information. The dissimilarities are in delta, the weights in weights. Dissimilarities are sorted in increasing order and the blocks vectors codes the tie-blocks in the dissimilarities.



## Initial Estimate

There are two possibilities. If xinit is not NULL then it is used as the 
initial estimate of the configuration. This is useful, for example,
if we start the iterations for, say, $r=\tfrac12$ with the solition for $r=1$.
If xinit is NULL then the function smacofSSRStressInit() computes an initial estimate using classical MDS (also known as Torgerson-Gower MDS) on the dissimilarities.

First we make sure the weights add up to one. If there are
missing dissimilarities they are first imputed by setting them equal to the 
weighted mean of the non-missing dissimilarities. Then we double center the squared
dissimilarities and compute the classical scaling solution using
the eigs_sym() function from the RSpectra package (@qiu_mei_24).

Now we scale the solution, which is done with the R function
smacofSSRStressScale(). We first normalize the powered dissimilarities
such that their weighted sum of squares is equal to one. We then
choose $\lambda$ to minimize
$$
\sigma(\lambda):=\sum_k w_k(\delta_k^r-d_k^r(\lambda x))^2=\sum_k w_k(\delta_k^r-\lambda^rd_k^r(x))^2,
$$
where $x$ is the Torgerson-Gower solution. The minimum is attained for
$$
\hat\lambda:=\left[\frac{\sum_k w_k\delta_k^rd_k^r(x)}{\sum_kw_kd_k^{2r}(x)}\right]^{1/r}.
$$
Our initial estimate is $\hat\lambda x$ with distances $\hat\lambda d_k(x)$. This scaling is done also if xinit is not NULL.

## Main

The arguments of fStress() are
```{r theargsf, echo = FALSE}
args(fStress)
```

Thus fStress() needs an initial configuration. Parameters $f$ and $g$ are functions that compute the transformation $f$ and its derivative. The file
fStress.R includes flog, glog, fexp, gexp, and fPower, gPower. Parameter
$r$ is included in case $f$ and $g$ need additional arguments (as in the case of 
fPower and gPower). Since all code is in R, it is easy to add additional 
transformation functions if needed. The output of fStress() is 
```{r outputf, eval = FALSE}
list(
    delta = delta, 
    dhat = delf, 
    weightmat = wght,
    conf = xnew,
    confdist = dnew,
    stress = snew,
    niter = itel
  )
```

The arguments of smacofSSRStress() are
```{r theargsr, echo = FALSE}
args(smacofSSRStress)
```
The C code follows the conventions of the smacof versions in @deleeuw_E_25b. 
There is only a single parameter $rpow$ for the transformation. The output
of smacofSSRStress() is
```{r output, eval = FALSE}
list(
    delta = theData$delta,
    dhat = h$dhat,
    confdist = h$edis,
    conf = matrix(h$xnew, nobj, ndim),
    weightmat = h$wght,
    stress = h$snew,
    ndim = ndim,
    init = xinit,
    niter = h$itel,
    nobj = nobj,
    iind = h$iind,
    jind = h$jind,
    weighted = weighted,
    ordinal = ordinal,
    ties = h$ties
  )
```
The output list can be fed immediately into the smacof plot routines for the
configuration plot, the residual plot, the Shepard plot, and the Dist-Dhat
plot.


# Examples

We analyse the colour data from @ekman_54 with different functions $f$, first 
using fStress(), then using smacofSSRStress().

```{r fstressruns, echo = FALSE, cache = FALSE}
hl <- fStress(ekmanData, f = flog, g = glog, verbose = FALSE)
hp1 <- fStress(ekmanData, f = fPower, g = gPower, r = .1, verbose = FALSE)
hp5 <- fStress(ekmanData, f = fPower, g = gPower, r = .5, verbose = FALSE)
h1 <- fStress(ekmanData, f = fPower, g = gPower, r = 1, verbose = FALSE)
h2 <- fStress(ekmanData, f = fPower, g = gPower, r = 2, verbose = FALSE)
h5 <- fStress(ekmanData, f = fPower, g = gPower, r = 5, verbose = FALSE)
he <- fStress(ekmanData, f = fexp, g = gexp, verbose = FALSE)
```

```{r kableF, echo = FALSE}
floss <- c(hl$stress, hp1$stress, hp5$stress, h1$stress,h2$stress,h5$stress,he$stress)
fiter <-  c(hl$niter, hp1$niter, hp5$niter, h1$niter,h2$niter,h5$niter,he$niter)
x <- matrix(c(fiter, floss), 7, 2)
r <- c("logarithm", 
       "power .1",
       "power .5",
       "power 1",
       "power 2", 
       "power 5",
       "exponent"
)
row.names(x) <- r
r <- c("niter", "fStress")
colnames(x) <- r
kable(x, format = "simple", align = NULL, row.names = TRUE, caption = "fStress() results Ekman Data")
```
```{r rstressruns, echo = FALSE, cache = FALSE}
chp1 <- smacofSSRStress(ekmanData, rpow = .1, verbose = FALSE)
chp5 <- smacofSSRStress(ekmanData, rpow = .5, verbose = FALSE)
ch1 <- smacofSSRStress(ekmanData, rpow = 1, verbose = FALSE)
ch2 <- smacofSSRStress(ekmanData, rpow = 2, verbose = FALSE)
ch5 <- smacofSSRStress(ekmanData, rpow = 5, verbose = FALSE)
```
```{r kableR, echo = FALSE}
rloss <- c(chp1$stress, chp5$stress, ch1$stress,ch2$stress,ch5$stress)
riter <-  c(chp1$niter, chp5$niter, ch1$niter,ch2$niter,ch5$niter)
x <- matrix(c(riter, rloss), 5, 2)
r <- c("power .1",
       "power .5",
       "power 1",
       "power 2", 
       "power 5"
)
row.names(x) <- r
r <- c("niter", "rStress")
colnames(x) <- r
kable(x, format = "simple", align = NULL, row.names = TRUE, caption = "smacofSSRStress() results Ekman Data, metric MDS")
```

We see that for the power transforms the R and C functions arrive at the
same solution. The number of iterations is not excatly the same, because
of differences in the initial configuration. With smacofSSRStress() we can also
do ordinal MDS.

```{r rstressrunsO, echo = FALSE, cache = FALSE}
chp1o <- smacofSSRStress(ekmanData, rpow = .1, verbose = FALSE, ordinal = TRUE)
chp5o <- smacofSSRStress(ekmanData, rpow = .5, verbose = FALSE, ordinal = TRUE)
ch1o <- smacofSSRStress(ekmanData, rpow = 1, verbose = FALSE, ordinal = TRUE)
ch2o <- smacofSSRStress(ekmanData, rpow = 2, verbose = FALSE, ordinal = TRUE)
ch5o <- smacofSSRStress(ekmanData, rpow = 5, verbose = FALSE, ordinal = TRUE)
```


```{r kableRO, echo = FALSE}
rlosso <- c(chp1o$stress, chp5o$stress, ch1o$stress,ch2o$stress,ch5o$stress)
ritero <-  c(chp1o$niter, chp5o$niter, ch1o$niter,ch2o$niter,ch5o$niter)
x <- matrix(c(ritero, rlosso), 5, 2)
r <- c("power .1",
       "power .5",
       "power 1",
       "power 2", 
       "power 5"
)
row.names(x) <- r
r <- c("niter", "rStress")
colnames(x) <- r
kable(x, format = "simple", align = NULL, row.names = TRUE, caption = "
      smacofSSRStress() results Ekman Data, ordinal MDS")
```

The package microbenchmark (@mersmann_24) can be used to compare the
R and C versions. For $r=1$ the C version is about four times faster,
for $r=.1$ and $r=5$ it is about 10 times faster.

```{r timer, echo = FALSE, cache = TRUE}
hmicro <- microbenchmark(
  fStress(ekmanMatrix, xini = xinit, f = fPower, g = gPower, r = 1, verbose = FALSE),
  smacofSSRStress(ekmanData, rpow = 1, verbose = FALSE),
  fStress(ekmanMatrix, xini = xinit, f = fPower, g = gPower, r = .1, verbose = FALSE),
  smacofSSRStress(ekmanData, rpow = .1, verbose = FALSE),
  fStress(ekmanMatrix, xini = xinit, f = fPower, g = gPower, r = 5, verbose = FALSE),
  smacofSSRStress(ekmanData, rpow = 5, verbose = FALSE))
print(hmicro)
```
\sectionbreak
```{r chp1c, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Metric, r = .1", main =""}
par(pty="s")
smacofConfigurationPlot(chp1, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r chp1s, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Metric, r = .1", main = ""}
smacofShepardPlot(chp1, rpow = .1, cex = 2)
```
\sectionbreak
```{r chp5c, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Metric, r = .5", main =""}
par(pty="s")
smacofConfigurationPlot(chp5, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r chp5s, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Metric, r = .5", main = ""}
smacofShepardPlot(chp5, rpow = .5, cex = 2)
```
\sectionbreak
```{r ch1c, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Metric, r = 1", main =""}
par(pty="s")
smacofConfigurationPlot(ch1, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r ch1s, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Metric, r = 1", main = ""}
smacofShepardPlot(ch1, rpow = 1, cex = 2)
```
\sectionbreak
```{r ch2c, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Metric, r = 2", main =""}
par(pty="s")
smacofConfigurationPlot(ch2, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r ch2s, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Metric, r = 2", main = ""}
smacofShepardPlot(ch2, rpow = 2, cex = 2)
```
\sectionbreak
```{r ch5c, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Metric, r = 5", main =""}
par(pty = "s")
smacofConfigurationPlot(ch5, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r ch5s, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Metric, r = 5", main = ""}
smacofShepardPlot(ch5, rpow = 5, cex = 2)
```

\sectionbreak
```{r chp1oc, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Non-metric, r = .1", main =""}
par(pty="s")
smacofConfigurationPlot(chp1o, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r chp1os, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Non-metric, r = .1", main = ""}
smacofShepardPlot(chp1o, rpow = 2, cex = 2)
```
\sectionbreak
```{r chp5oc, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Non-metric, r = .5", main =""}
par(pty="s")
smacofConfigurationPlot(chp5o, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r chp5os, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Non-metric, r = .5", main = ""}
smacofShepardPlot(chp5o, rpow = .5, cex = 2)
```
\sectionbreak
```{r ch1oc, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Non-metric, r = 1", main =""}
par(pty = "s")
smacofConfigurationPlot(ch1o, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r ch1os, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Non-metric, r = 1", main = ""}
smacofShepardPlot(ch1o, rpow = 1, cex = 2)
```
\sectionbreak
```{r ch2oc, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Non-metric, r = 2", main =""}
par(pty = "s")
smacofConfigurationPlot(ch2o, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r ch2os, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Non-metric, r = 2", main = ""}
smacofShepardPlot(ch2o, rpow = 2, cex = 2)
```
\sectionbreak
```{r ch5oc, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Configuration Plot, Non-metric, r = 5", main =""}
par(pty = "s")
smacofConfigurationPlot(ch5o, labels = ekmanLabels, cex = 2)
```
\sectionbreak
```{r ch5os, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Shepard Plot, Non-metric, r = 5", main = ""}
smacofShepardPlot(ch5o, rpow = 5, cex = 2)
```



\sectionbreak

# References