% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{241,243,245}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother


% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}



\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}



 


\usepackage{tcolorbox}
\usepackage{amssymb}
\usepackage{yfonts}
\usepackage{bm}


\newtcolorbox{greybox}{
  colback=white,
  colframe=blue,
  coltext=black,
  boxsep=5pt,
  arc=4pt}
  
\newcommand{\sectionbreak}{\clearpage}

 
\newcommand{\ds}[4]{\sum_{{#1}=1}^{#3}\sum_{{#2}=1}^{#4}}
\newcommand{\us}[3]{\mathop{\sum\sum}_{1\leq{#2}<{#1}\leq{#3}}}

\newcommand{\ol}[1]{\overline{#1}}
\newcommand{\ul}[1]{\underline{#1}}

\newcommand{\amin}[1]{\mathop{\text{argmin}}_{#1}}
\newcommand{\amax}[1]{\mathop{\text{argmax}}_{#1}}

\newcommand{\ci}{\perp\!\!\!\perp}

\newcommand{\mc}[1]{\mathcal{#1}}
\newcommand{\mb}[1]{\mathbb{#1}}
\newcommand{\mf}[1]{\mathfrak{#1}}

\newcommand{\eps}{\epsilon}
\newcommand{\lbd}{\lambda}
\newcommand{\alp}{\alpha}
\newcommand{\df}{=:}
\newcommand{\am}[1]{\mathop{\text{argmin}}_{#1}}
\newcommand{\ls}[2]{\mathop{\sum\sum}_{#1}^{#2}}
\newcommand{\ijs}{\mathop{\sum\sum}_{1\leq i<j\leq n}}
\newcommand{\jis}{\mathop{\sum\sum}_{1\leq j<i\leq n}}
\newcommand{\sij}{\sum_{i=1}^n\sum_{j=1}^n}
	
\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Minimizing fStress and rStress by Majorizing Gauss-Newton},
  pdfauthor={Jan de Leeuw},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}


\title{Minimizing fStress and rStress by Majorizing Gauss-Newton}
\author{Jan de Leeuw}
\date{March 5, 2026}
\begin{document}
\maketitle
\begin{abstract}
TBD
\end{abstract}

\renewcommand*\contentsname{Table of contents}
{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{3}
\tableofcontents
}

\sectionbreak

\textbf{Note:} This is a working manuscript which will be
expanded/updated frequently. All suggestions for improvement are
welcome. All Rmd, tex, html, pdf, R, and C files are in the public
domain. Attribution will be appreciated, but is not required. The files
can be found at \url{https://github.com/deleeuw/rStress}

\sectionbreak

\section{Loss Functions}\label{loss-functions}

The Multidimensional Scaling (MDS) loss function fStress is defined as
\begin{equation}
\sigma_f(x):=\frac12\sum_{k=1}^K w_k(f(\delta_k)-f(d_k(x)))^2,\label{eq-fdef}
\end{equation} with \(f\) increasing and differentiable in the open
interval \((0,+\infty)\). In \eqref{eq-fdef} the \(w_k\) are positive
\emph{weights}, the \(\delta_k\) are known \emph{dissimilarities}. The
vector \(x\) has the coordinates of \(n\) points in \(\mathbb{R}^p\),
and the \(d_k(x)\) are Euclidean \emph{distances} between pairs of these
points. For each \(k\) there is a pair of indices \((i,j)\) with
\(1\leq i<j\leq n\) and a matrix \(A_k\) which is the direct sum of
\(p\) copies of \((e_i-e_j)(e_i-e_j)'\), where the \(e_i\) are unit
vectors (columns of the identity matrix of order \(n\)). Thus
\(\smash{d_k(x)=\sqrt{x'A_kx}}\). Metric least squares MDS minimizes
fStress over \(x\).

fStress was first introduced and studied in Groenen, De Leeuw, and
Mathar (\citeproc{ref-groenen_deleeuw_mathar_C_95}{1995}). No explicit
algorithm to minimize loss \eqref(eq-fdef) was given, but the paper has
formulas for the first and second derivatives (see also De Leeuw
(\citeproc{ref-deleeuw_E_17r}{2017})), which can of course be used in
general purpose optimization routines.

An important special case of fStress is rStress (also known as
powerStress), which is \begin{equation}
\sigma_r(x):=\frac12\sum_{k=1}^K w_k(\delta_k^r-d_k^r(x))^2\label{eq-rdef}
\end{equation} Special cases of rStress are Kruskal's stress (Kruskal
(\citeproc{ref-kruskal_64a}{1964a}), Kruskal
(\citeproc{ref-kruskal_64b}{1964b})) with \(r=1\), sstress by Takane,
Young, and De Leeuw (\citeproc{ref-takane_young_deleeuw_A_77}{1977})
with \(r=2\), and the limiting case logarithmic stress with
\(r\rightarrow 0\) by Ramsay (\citeproc{ref-ramsay_77}{1977}). There
have been various attempts to extend the majorization (or MM) method for
MDS (De Leeuw (\citeproc{ref-deleeuw_C_77}{1977})) to rStress.
References and links to various unpublished reports are in De Leeuw
(\citeproc{ref-deleeuw_E_17r}{2017}). The recent smacofx package (Rusch
et al. (\citeproc{ref-rusch_deleeuw_mair_hornik_25}{In Press})) has R
code for the rstressMin() function that implements the majorization
method in De Leeuw, Groenen, and Mair
(\citeproc{ref-deleeuw_groenen_mair_E_16a}{2016}).

Minimizing of either fStress or rStress over \(x\) is a metric MDS
problem. The \(\delta_k\), and consequently the \(f(\delta_k)\), are
\(K\) known numbers. In non-metric MDS the loss function is minimized
over both \(x\) and \(\delta\), where \(\delta\) is constrained to be in
a subset \(\Delta\) of \(\mathbb{R}^K\). For the ordinal version of
non-metric MDS, for example, we require
\(\delta_1\leq\cdots\leq\delta_K\). Since \(f\) is increasing, we can
write fStress simply as \begin{equation}
\sigma_f(x,\delta):=\frac12\sum_{k=1}^K w_k(\delta_k-f(d_k(x)))^2,\label{eq-nmfdef}
\end{equation} where \(\delta\) is no longer a vector of known
dissimilarities, but any vector monotone with the dissimilarities. These
transformed or scaled dissimilarities are often called
\emph{disparities}. Non-metric rStress is simply \eqref{eq-nmfdef} with
\(\smash{f(d_k(x))=d_k^r(x)}\).

In order to exclude the trivial solution with \(x=0\) and \(\delta=0\)
in addition we impose the normalization constraint \begin{equation}
\eta^2(\delta):=\frac12\sum_{k=1}^Kw_k\delta_k^2=1.\label{eq-normal}
\end{equation} This formulation of non-metric MDS can be generalized to
\(\delta\in\Delta\), where \(\Delta\) is a convex cone in
\(\mathbb{R}^K\). This means we require the disparities to be in the
intersection of the cone \(\Delta\) and the sphere \(\Sigma\) defined by
\eqref{eq-normal}.

It is convenient to think of metric MDS as the special case in which the
set \(\Delta\) is the one-element set containing only the normalized
dissimilarities \(\delta\). Alternatively, as in smacof, we can choose
for \(\Delta\) the ray of all vectors that are non-negative multiples of
the dissimilarities. In other words our metric MDS treats the
dissimilarities as measured on a ratio scale.

\section{Algorithm}\label{algorithm}

The technique proposed in this paper to minimize non-metric fStress or
rStress is in the Alternating Least Squares (ALS) family. We start with
an initial estimate \(x^{(0)}\). We then minimize \(\sigma_f\) over
\(\delta\) in \(\Delta\cap\Sigma\) for the current
\(\smash{d(x^{(0)})}\). The minimizer \(\smash{\delta^{(0)}}\) is then
used to minimize fStress over \(x\) for the current disparities,
yielding \(x^{(1)}\). These two steps are alternated until \(x\) and
\(\delta\) do not change any more. Starting in iteration \(\nu=0\) we
compute \begin{subequations}
\begin{align}
\delta^{(\nu)}&=\mathop{\text{argmin}}_{\delta\in\Delta\cap\Sigma}\sigma_f(x^{(\nu)},\delta),\label{eq-als1}\\
x^{(\nu+1)}&=\mathop{\text{argmin}}_x\sigma_f(x,\delta^{(\nu)})\label{eq-als2}.
\end{align}
\end{subequations} We then increase \(\nu\) by one and go into the next
iteration. And so on, until convergence.

\subsection{Normalized Cone
Regression}\label{normalized-cone-regression}

In metric MDS there is no need for the first step \eqref{eq-als1},
because \(\delta^{(\nu)}\) is equal to the normalized dissimilarities
\(\delta\) for all \(\nu\). In the non-metric case step \eqref{eq-als1}
is needed, but compared to \eqref{eq-als2} it is comparatively easy. We
have to compute the least squares projection of \(f(d(x))\) on the cone
\(\Delta\) and then normalize this projection to give it length one. If
\(\Delta\) is the cone of monotone vectors, as in ordinal MDS, the cone
projection is \emph{monotone regression}. The fact that projection on
the intersection of \(\Delta\) and \(\Sigma\) is the same thing as
normalizing the projection on \(\Delta\) is due to De Leeuw
(\citeproc{ref-deleeuw_U_75a}{1975}) and more generally (and more
rigorously) to Bauschke, Bui, and Wang
(\citeproc{ref-bauschke_bui_wang_18}{2018}). Since this step is the same
as in standard MDS algorithms such as smacof (De Leeuw and Mair
(\citeproc{ref-deleeuw_mair_A_09c}{2009}), Mair, Groenen, and De Leeuw
(\citeproc{ref-mair_groenen_deleeuw_A_22}{2022})) we do not go into
details here, and just refer to the literature.

\subsection{Majorization}\label{majorization}

The second step \eqref{eq-als2}, minimizing over \(x\) for fixed current
\(\delta\), is more complicated. There is no analytic solution, unlike
the first step, and minimization requires an iterative process of its
own. Thus, except in some very special cases, the second step requires
an infinite number of ``inner'' iterations. Since implementing an
infinite number of iterations is impossible we have to truncate the
inner iteration sequence at some point. In our software we deviate from
the strict ALS framework by not minimizing fStress over \(x\) for fixed
\(\delta\), but by taking a single ``inner'' iteration and merely
decrease fStress. If this is done judiciously we still obtain a
convergent sequence of updates. This is similar to the strategy in other
MDS algorithms such as smacof (De Leeuw
(\citeproc{ref-deleeuw_C_77}{1977})) and alscal (Takane, Young, and De
Leeuw (\citeproc{ref-takane_young_deleeuw_A_77}{1977})).

In smacof the inner iteration step is a majorization step, by now more
commonly known as an MM step. Briefly, we find a function
\(\kappa_f(x;y)\) such that

\begin{itemize}
\tightlist
\item
  \(\kappa_f(x,y)\leq\sigma_f(x)\) for all \(x\) and \(y\) in \(\Xi\),
  and
\item
  \(\kappa_f(x;y)=\sigma_f(x)\) if and only if \(x=y\).
\end{itemize}

An inner iteration is of the form \begin{equation}
x^{(\mu+1)}=\mathop{\text{argmin}}_{x\in\Xi}\kappa(x,x^{(\mu)}).\label{eq-inner}
\end{equation} If \(x^{(\mu+1)}=x^{(\mu)}\) we declare convergence and
stop. From \eqref{eq-inner}, \begin{equation}
\sigma_f(x^{(\mu+1)})\leq\kappa_f(x^{(\mu+1)},x^{(\mu)})<\kappa_f(x^{(\mu)},x^{(\mu)})=\sigma_f(x^{(\mu)}).
\label{eq-sandwich}
\end{equation} Thus an inner majorization step upgrading \(x\) for given
\(\delta\) decreases fStress (unless we stop). Remember that in ALS the
inner iterative process (indexed by \(\mu\)) goes on within step 2 of
the ``outer'' update (indexed by \(\nu\)).

In De Leeuw, Groenen, and Mair
(\citeproc{ref-deleeuw_groenen_mair_E_16a}{2016}) a majorization method
was proposed to minimize rStress. It majorizes \(d^r\) as a function of
\(x\), and for this it uses the specific properties of the power
function. It consequently cannot be used for fStress. The technique is
incorporated in the smacofx R package, and numerical experience
indicates it works, but convergence can be painfully slow. Because of
the generality of the function \(f\) (any differentiable increasing
function) it seems impossible to develop an majorization method for
fStress along the same lines as the one for rStress. Consequently we go
a somewhat different route in this paper. We do not only deviate from
the strict ALS procedure by only partially minimizing loss over \(x\)
for fixed \(d\), we also deviate from the strict majorization approach
by majorizing an approximation of fStress.

In deriving the approximation and majorization we surpress the
dependency of fStress on \(\delta\), because in the second ALS substep
\(\delta\) is just a vector of constants. We first approximate
\(f(d(x))\) near \(d(y)\) with \begin{equation}
f(d_k(x))\approx f(d_k(y))+\mathcal{D}f(d_k(y))(d_k(x)-d_k(y)).\label{eq-taylor}
\end{equation} Define \begin{equation}
\omega_f(x;y):=\sum_{k=1}^K w_k(f(\delta_k)-f(d_k(y))-\mathcal{D}f(d_k(y))(d_k(x)-d_k(y)))^2,\label{eq-defeta}
\end{equation} Note that in \eqref{eq-taylor} the derivative is with
respect to \(d\), not with respect to \(x\) or \(y\). This is a major
difference with the majorizations in De Leeuw, Groenen, and Mair
(\citeproc{ref-deleeuw_groenen_mair_E_16a}{2016}).

Note that \(\omega_f(x;x)=\sigma_f(x)\) for all \(x\) and if \(f\) is
the identity then \(\omega_f(x;y)=\sigma_f(x)\) for all \(x\) and \(y\).
If \(f\) is linear with intercept \(\beta\) and slope \(\alpha\) then
\begin{equation}
\omega_f(x,y)=a^2\sum_{k=1}^Kw_k(\delta_k-d_k(x))^2=a^2\sigma_r(x),\label{eq-linear}
\end{equation} with \(r=1\).

We now give an alternative and more convenient expression for
\(\omega_f(x,y)\). Define \begin{subequations}
\begin{align}
\tilde w_k(y)&:=w_k\{\mathcal{D}f(d_k(y))\}^2,\label{eq-tildew}\\
\tilde\delta_k(y)&:=\frac{f(\delta_k)-f(d_k(y))}{\mathcal{D}f(d_k(y))}+d_k(y).\label{eq-tilded}
\end{align}
\end{subequations} Then \begin{equation}
\omega_f(x;y)=\sum_{k=1}^K\tilde w_k(y)(\tilde\delta_k(y)-d_k(x))^2.\label{eq-omegas}
\end{equation} We see from \eqref{eq-tildew} that \(\tilde w_k(y)\) is
always non-negative and from \eqref{eq-tilded} that
\(\tilde\delta_k(y)\) can be negative if \(d_k(y)>\delta_k\). For convex
\(f\), however, we have \begin{equation}
\tilde\delta_k(y)=\frac{f(\delta_k)-f(d_k(y))}{\mathcal{D}f(d_k(y))}+d_k(y)\geq d_k(x)\geq 0.\label{eq-convex}
\end{equation}

For rStress equations \eqref{eq-tildew} and \eqref{eq-tilded} become
\begin{subequations}
\begin{align}
\tilde w_k(y)&:=r^2w_kd_k^{2(r-1)}(y),\label{eq-tildewr}\\
\tilde\delta_k(y)&:=\frac{\delta_k^r+(r-1)d_k^r(y)}{rd_k^{r-1}(y)}.\label{eq-tildedr}
\end{align}
\end{subequations} By convexity \(\tilde\delta_k(y)\) is non-negative if
\(r\geq 1\). For \(0<r<1\) we have \(\tilde\delta_k(y)>0\) if and only
if \(\smash{d_k(y)<(1-r)^r\delta_k}\). If \(\delta_k>d_k(y)\) then
\(\tilde\delta_k(y)>d_k(y)\geq 0\).

By writing \(\omega_f(x,y)\) as in \eqref{eq-omegas} we are on familiar
majorization terrain. If \(\delta_k(y)>0\) we use Cauchy-Schwartz, as in
standard smacof, \begin{subequations}
\begin{equation}
d_k(x)\geq\frac{1}{d_k(y)}\text{tr}\ x'A_ky,\label{eq-cs}
\end{equation}
and if $\delta_k(y)<0$ we use the arithmetic mean/geometric mean (AM/GM) inequality
in the form
\begin{equation}
d_k(x)\leq\frac12\frac{1}{d_k(y)}\{y'A_ky+x'A_kx\},\label{eq-amgm}
\end{equation}
\end{subequations} Use of the AM/GM inequality for majorizing terms with
negative dissimilarities was pioneered by Heiser
(\citeproc{ref-heiser_91}{1991}).

Define the matrices \begin{subequations}
\begin{align}
V(y)&:=\sum_{k=1}^K\tilde w_k(y)A_k,\\
B(y)&:=\sum_{\tilde\delta_k(y)>0}\tilde w_k\frac{\tilde\delta_k(y)}{d_k(y)}A_k,\\
H(y)&:=\sum_{\tilde\delta_k(y)<0}\tilde w_k\frac{\tilde\delta_k(y)}{d_k(y)}A_k.
\end{align}
\end{subequations} Note that all three matrices (more preecisely
matrix-valued functions) are positive semi-definite. We can use these
matrices for a majorization of \(\omega_f(x;y)\) at \(y\) is
\begin{equation}
\omega_f(x;y)=C+\frac12 x'(V(y)+H(y))x-x'B(y)x\leq\kappa_f(x;y),\label{eq-matmajx}
\end{equation} with \begin{equation}
\kappa_f(x;y):=C+\frac12 x'(V(y)+H(y))x-x'B(y)y,\label{eq-matmajy}
\end{equation} where \(C\) is a constant that depends on \(y\) but not
on \(x\).

We now update using \[
x^{(\mu+1)}=\mathop{\text{argmin}}_x\kappa_f(x;x^{(\mu)})=(V(x^{(\mu)})+H(x^{(\mu)}))^+B(x^{(\mu)})x^{(\mu)},
\] which results in \[
\sigma_f(x^{(\mu+1)})\approx\omega_f(x^{(\mu+1)};x^{(\mu)})\leq\kappa_f(x^{(\mu+1)};x^{(\mu)})<\kappa_f(x^{(\mu)};x^{(\mu)})=\sigma_f(x^{(\mu)}).
\] Unlike in smacof there is no guarantee that \(\sigma_f\) decreases
from one iteration to the other. Monotone convergence of loss function
values will only happen if the approximation is good, i.e.~if
\(d(x^{(\mu+1)})\) is close to \(d_k(x^{(\mu)})\). This requires closer
monitoring of convergence than in smacof. Also note that if
\(\tilde\delta_k(y)<0\) for all \(k\) then \(B(y)=0\), which means that
the update from \(y\) is equal to zero. Again, this eventuality must be
monitored.

\section{Software}\label{software}

There are two different major routines in the repository. The R function
fStress.R does metric MDS minimizing fStress with an arbitrary \(f\).
The C function smacofSSRStress.c does either metric or ordinal MDS
minimizing rStress for an arabitrary \(r\). The C code is compiled into
a shared library smacofSSRStress.so, which is loaded by the R front-end
smacofSSRStress.R that contains smacofSSRStress(). The function
fStress() is meant for exploration and research, while smacofSSRStress()
is more suitable for production. fStress() is more general, in that it
allows for more transformations, but smacofSSRStress() is more general
in that it allows for ordinal MDS.

\subsection{Data}\label{data}

For fStress.R the input is a matrix delta of dissimilarities. For
smacofSSRStress.R the data and weights are in the MDS data format
defined in De Leeuw (\citeproc{ref-deleeuw_E_25b}{2025}). This is a list
with two scalars nobj and ndat, the number of objects and the number of
data points (object pairs), and five vectors iind, jind, delta, blocks,
weights of length ndat. Vectors iind and jind have integers between 1
and n with iind \textgreater{} jind that code for which pairs of objects
we have dissimilarity information. The dissimilarities are in delta, the
weights in weights. Dissimilarities are sorted in increasing order and
the blocks vectors codes the tie-blocks in the dissimilarities.

\subsection{Initial Estimate}\label{initial-estimate}

There are two possibilities. If xinit is not NULL then it is used as the
initial estimate of the configuration. This is useful, for example, if
we start the iterations for, say, \(r=\tfrac12\) with the solition for
\(r=1\). If xinit is NULL then the function smacofSSRStressInit()
computes an initial estimate using classical MDS (also known as
Torgerson-Gower MDS) on the dissimilarities.

First we make sure the weights add up to one. If there are missing
dissimilarities they are first imputed by setting them equal to the
weighted mean of the non-missing dissimilarities. Then we double center
the squared dissimilarities and compute the classical scaling solution
using the eigs\_sym() function from the RSpectra package (Qiu and Mei
(\citeproc{ref-qiu_mei_24}{2024})).

Now we scale the solution, which is done with the R function
smacofSSRStressScale(). We first normalize the powered dissimilarities
such that their weighted sum of squares is equal to one. We then choose
\(\lambda\) to minimize \[
\sigma(\lambda):=\sum_k w_k(\delta_k^r-d_k^r(\lambda x))^2=\sum_k w_k(\delta_k^r-\lambda^rd_k^r(x))^2,
\] where \(x\) is the Torgerson-Gower solution. The minimum is attained
for \[
\hat\lambda:=\left[\frac{\sum_k w_k\delta_k^rd_k^r(x)}{\sum_kw_kd_k^{2r}(x)}\right]^{1/r}.
\] Our initial estimate is \(\hat\lambda x\) with distances
\(\hat\lambda d_k(x)\). This scaling is done also if xinit is not NULL.

\subsection{Main}\label{main}

The arguments of fStress() are

\begin{verbatim}
function (delta, xini, wght = 1 - diag(nrow(xini)), f = fPower, 
    g = gPower, r = 1, itmax = 10000, eps = 1e-06, verbose = TRUE) 
NULL
\end{verbatim}

Thus fStress() needs an initial configuration. Parameters \(f\) and
\(g\) are functions that compute the transformation \(f\) and its
derivative. The file fStress.R includes flog, glog, fexp, gexp, and
fPower, gPower. Parameter \(r\) is included in case \(f\) and \(g\) need
additional arguments (as in the case of fPower and gPower). Since all
code is in R, it is easy to add additional transformation functions if
needed. The output of fStress() is

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{list}\NormalTok{(}
    \AttributeTok{delta =}\NormalTok{ delta, }
    \AttributeTok{dhat =}\NormalTok{ delf, }
    \AttributeTok{weightmat =}\NormalTok{ wght,}
    \AttributeTok{conf =}\NormalTok{ xnew,}
    \AttributeTok{confdist =}\NormalTok{ dnew,}
    \AttributeTok{stress =}\NormalTok{ snew,}
    \AttributeTok{niter =}\NormalTok{ itel}
\NormalTok{  )}
\end{Highlighting}
\end{Shaded}

The arguments of smacofSSRStress() are

\begin{verbatim}
function (theData, ndim = 2, xinit = NULL, ties = 1, itmax = 10000, 
    eps = 1e-06, rpow = 1, digits = 8, width = 10, verbose = TRUE, 
    weighted = FALSE, ordinal = FALSE) 
NULL
\end{verbatim}

The C code follows the conventions of the smacof versions in De Leeuw
(\citeproc{ref-deleeuw_E_25b}{2025}). There is only a single parameter
\(rpow\) for the transformation. The output of smacofSSRStress() is

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{list}\NormalTok{(}
    \AttributeTok{delta =}\NormalTok{ theData}\SpecialCharTok{$}\NormalTok{delta,}
    \AttributeTok{dhat =}\NormalTok{ h}\SpecialCharTok{$}\NormalTok{dhat,}
    \AttributeTok{confdist =}\NormalTok{ h}\SpecialCharTok{$}\NormalTok{edis,}
    \AttributeTok{conf =} \FunctionTok{matrix}\NormalTok{(h}\SpecialCharTok{$}\NormalTok{xnew, nobj, ndim),}
    \AttributeTok{weightmat =}\NormalTok{ h}\SpecialCharTok{$}\NormalTok{wght,}
    \AttributeTok{stress =}\NormalTok{ h}\SpecialCharTok{$}\NormalTok{snew,}
    \AttributeTok{ndim =}\NormalTok{ ndim,}
    \AttributeTok{init =}\NormalTok{ xinit,}
    \AttributeTok{niter =}\NormalTok{ h}\SpecialCharTok{$}\NormalTok{itel,}
    \AttributeTok{nobj =}\NormalTok{ nobj,}
    \AttributeTok{iind =}\NormalTok{ h}\SpecialCharTok{$}\NormalTok{iind,}
    \AttributeTok{jind =}\NormalTok{ h}\SpecialCharTok{$}\NormalTok{jind,}
    \AttributeTok{weighted =}\NormalTok{ weighted,}
    \AttributeTok{ordinal =}\NormalTok{ ordinal,}
    \AttributeTok{ties =}\NormalTok{ h}\SpecialCharTok{$}\NormalTok{ties}
\NormalTok{  )}
\end{Highlighting}
\end{Shaded}

The output list can be fed immediately into the smacof plot routines for
the configuration plot, the residual plot, the Shepard plot, and the
Dist-Dhat plot.

\section{Examples}\label{examples}

We analyse the colour data from Ekman (\citeproc{ref-ekman_54}{1954})
with different functions \(f\), first using fStress(), then using
smacofSSRStress().

\begin{longtable}[]{@{}lcc@{}}
\caption{fStress() results Ekman Data}\tabularnewline
\toprule\noalign{}
& niter & fStress \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
& niter & fStress \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
logarithm & 555 & 0.0872002 \\
power .1 & 805 & 0.0002733 \\
power .5 & 219 & 0.0059543 \\
power 1 & 35 & 0.0172132 \\
power 2 & 45 & 0.0328807 \\
power 5 & 775 & 0.0203153 \\
exponent & 34 & 0.0085331 \\
\end{longtable}

\begin{longtable}[]{@{}lcc@{}}
\caption{smacofSSRStress() results Ekman Data, metric
MDS}\tabularnewline
\toprule\noalign{}
& niter & rStress \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
& niter & rStress \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
power .1 & 856 & 0.0002733 \\
power .5 & 229 & 0.0059543 \\
power 1 & 35 & 0.0172132 \\
power 2 & 46 & 0.0328807 \\
power 5 & 841 & 0.0203153 \\
\end{longtable}

We see that for the power transforms the R and C functions arrive at the
same solution. The number of iterations is not excatly the same, because
of differences in the initial configuration. With smacofSSRStress() we
can also do ordinal MDS.

\begin{longtable}[]{@{}lcc@{}}
\caption{smacofSSRStress() results Ekman Data, ordinal
MDS}\tabularnewline
\toprule\noalign{}
& niter & rStress \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
& niter & rStress \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
power .1 & 876 & 0.0000066 \\
power .5 & 251 & 0.0001651 \\
power 1 & 134 & 0.0005337 \\
power 2 & 209 & 0.0009015 \\
power 5 & 2443 & 0.0011002 \\
\end{longtable}

The package microbenchmark (Mersmann (\citeproc{ref-mersmann_24}{2024}))
can be used to compare the R and C versions. For \(r=1\) the C version
is about four times faster, for \(r=.1\) and \(r=5\) it is about 10
times faster.

\begin{verbatim}
Warning in microbenchmark(fStress(ekmanMatrix, xini = xinit, f = fPower, : less
accurate nanosecond times to avoid potential integer overflows
\end{verbatim}

\begin{verbatim}
Unit: microseconds
                                                                                      expr
   fStress(ekmanMatrix, xini = xinit, f = fPower, g = gPower, r = 1,      verbose = FALSE)
                                     smacofSSRStress(ekmanData, rpow = 1, verbose = FALSE)
 fStress(ekmanMatrix, xini = xinit, f = fPower, g = gPower, r = 0.1,      verbose = FALSE)
                                   smacofSSRStress(ekmanData, rpow = 0.1, verbose = FALSE)
   fStress(ekmanMatrix, xini = xinit, f = fPower, g = gPower, r = 5,      verbose = FALSE)
                                     smacofSSRStress(ekmanData, rpow = 5, verbose = FALSE)
       min        lq      mean    median        uq       max neval cld
  1336.641  1429.445  1731.335  1470.588  1518.722  5500.847   100 a  
   363.875   389.377   439.438   412.419   484.743   723.199   100 a  
 33871.412 37650.772 45543.970 38470.259 39949.887 83167.475   100  b 
  3383.894  3611.997  3654.865  3653.530  3702.013  3955.844   100 a  
 32197.628 35727.543 39952.617 36587.826 37245.897 80553.807   100   c
  3288.323  3471.900  3525.207  3510.092  3582.129  3783.029   100 a  
\end{verbatim}

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-bauschke_bui_wang_18}
Bauschke, Heinz H., Minh N. Bui, and Xianfu Wang. 2018. {``{Projecting
onto the Intersection of a Cone and a Sphere}.''} \emph{SIAM Journal on
Optimization} 28: 2158--88.

\bibitem[\citeproctext]{ref-deleeuw_U_75a}
De Leeuw, Jan. 1975. {``{A Normalized Cone Regression Approach to
Alternating Least Squares Algorithms}.''} Department of Data Theory
FSW/RUL.
\url{https://jansweb.netlify.app/publication/deleeuw-u-75-a/deleeuw-u-75-a.pdf}.

\bibitem[\citeproctext]{ref-deleeuw_C_77}
---------. 1977. {``Applications of Convex Analysis to Multidimensional
Scaling.''} In \emph{Recent Developments in Statistics}, edited by J. R.
Barra, F. Brodeau, G. Romier, and B. Van Cutsem, 133--45. Amsterdam, The
Netherlands: North Holland Publishing Company.

\bibitem[\citeproctext]{ref-deleeuw_E_17r}
---------. 2017. {``{Higher Partials of fStress. Who Needs Them ?}''}
2017.
\url{https://jansweb.netlify.app/publication/deleeuw-e-17-r/deleeuw-e-17-r.pdf}.

\bibitem[\citeproctext]{ref-deleeuw_E_25b}
---------. 2025. {``Yet Another Smacof - Square Symmetric Case.''} 2025.
\url{https://jansweb.netlify.app/publication/deleeuw-e-25-b/}.

\bibitem[\citeproctext]{ref-deleeuw_groenen_mair_E_16a}
De Leeuw, Jan, Patrick Groenen, and Patrick Mair. 2016. {``{Minimizing
rStress Using Majorization}.''} 2016.
\url{https://jansweb.netlify.app/publication/deleeuw-groenen-mair-e-16-a/deleeuw-groenen-mair-e-16-a.pdf}.

\bibitem[\citeproctext]{ref-deleeuw_mair_A_09c}
De Leeuw, Jan, and Patrick Mair. 2009. {``{Multidimensional Scaling
Using Majorization: SMACOF in R}.''} \emph{Journal of Statistical
Software} 31 (3): 1--30.
\url{https://www.jstatsoft.org/article/view/v031i03}.

\bibitem[\citeproctext]{ref-ekman_54}
Ekman, Gosta. 1954. {``{Dimensions of Color Vision}.''} \emph{Journal of
Psychology} 38: 467--74.

\bibitem[\citeproctext]{ref-groenen_deleeuw_mathar_C_95}
Groenen, Patrick J. F., Jan De Leeuw, and Rudolf Mathar. 1995. {``{Least
Squares Multidimensional Scaling with Transformed Distances}.''} In
\emph{{From Data to Knowledge: Theoretical and Practical Aspects of
Classification, Data Analysis and Knowledge Organization}}, edited by W.
Gaul and D. Pfeifer. Berlin, Germany: Springer Verlag.
\url{https://jansweb.netlify.app/publication/groenen-deleeuw-mathar-c-95/groenen-deleeuw-mathar-c-95.pdf}.

\bibitem[\citeproctext]{ref-heiser_91}
Heiser, Willem J. 1991. {``{A Generalized Majorization Method for Least
Squares Multidimensional Scaling of Pseudodistances that May Be
Negative}.''} \emph{Psychometrika} 56 (1): 7--27.

\bibitem[\citeproctext]{ref-kruskal_64a}
Kruskal, Joseph B. 1964a. {``{Multidimensional Scaling by Optimizing
Goodness of Fit to a Nonmetric Hypothesis}.''} \emph{Psychometrika} 29:
1--27.

\bibitem[\citeproctext]{ref-kruskal_64b}
---------. 1964b. {``{Nonmetric Multidimensional Scaling: a Numerical
Method}.''} \emph{Psychometrika} 29: 115--29.

\bibitem[\citeproctext]{ref-mair_groenen_deleeuw_A_22}
Mair, Patrick, Patrick J. F. Groenen, and Jan De Leeuw. 2022. {``{More
on Multidimensional Scaling in R: smacof Version 2}.''} \emph{Journal of
Statistical Software} 102 (10): 1--47.
\url{https://www.jstatsoft.org/article/view/v102i10}.

\bibitem[\citeproctext]{ref-mersmann_24}
Mersmann, Olaf. 2024. \emph{{microbenchmark: Accurate Timing
Functions}}. \url{https://CRAN.R-project.org/package=microbenchmark}.

\bibitem[\citeproctext]{ref-qiu_mei_24}
Qiu, Yixuan, and Jiali Mei. 2024. \emph{{RSpectra: Solvers for
Large-Scale Eigenvalue and SVD Problems}}.
\url{https://CRAN.R-project.org/package=RSpectra}.

\bibitem[\citeproctext]{ref-ramsay_77}
Ramsay, James O. 1977. {``{Maximum Likelihood Estimation in
Multidimensional Scaling}.''} \emph{Psychometrika} 42: 241--66.

\bibitem[\citeproctext]{ref-rusch_deleeuw_mair_hornik_25}
Rusch, Thomas, Jan De Leeuw, Patrick Mair, and Kurt Hornik. In Press.
{``Flexible Multidimensional Scaling with the r Package Smacofx.''}
\emph{Journal of Statistical Software}, In Press.
\url{https://jansweb.netlify.app/publication/rusch-deleeuw-mair-hornik-a-25/rusch-deleeuw-mair-hornik-a-25.pdf}.

\bibitem[\citeproctext]{ref-takane_young_deleeuw_A_77}
Takane, Yoshio, Forrest W. Young, and Jan De Leeuw. 1977. {``Nonmetric
Individual Differences in Multidimensional Scaling: An Alternating Least
Squares Method with Optimal Scaling Features.''} \emph{Psychometrika}
42: 7--67.

\end{CSLReferences}




\end{document}
